---
title: "Hands-on Exercise 1"
subtitle: "A Layered Grammar of Graphics: ggplot2 methods"
author: "Thi Thanh Van Tran"
date: "April 16, 2026"
date-modified: "last-modified"
format:
  html:
    toc: true
    number-sections: true
    number-depth: 3
    code-fold: false
    code-tools: true
    code-copy: true
execute:
  eval: true
  echo: true
  warning: false
  freeze: false
---

**Learning Objectives:**

-   Master basic principles and essential components of ggplot2
-   Plot statistical graphics based on the principle of Layered Grammar of Graphics
-   Conduct data-driven analysis with dynamic insights
-   Extract meaningful patterns from visualizations

# Learning Outcome

In this chapter, I have learned the basic principles and essential components of ggplot2. At the same time, I've gained hands-on experience using these components to plot statistical graphics based on the principle of Layered Grammar of Graphics. Now I am able to apply the essential graphical elements provided by ggplot2 to create elegant and functional statistical graphics while extracting meaningful insights from data.

# Getting Started

## Installing and Loading Required Libraries

::: callout-note
**Note:** Ensure that the [pacman](https://cran.r-project.org/web/packages/pacman/) package has already been installed.
:::

The code chunk below uses `p_load()` of pacman package to check if tidyverse packages are installed in the computer. If they are, then they will be launched into R. Otherwise, tidyverse will be installed and launched into R.

```{r}
pacman::p_load(tidyverse)
```

## Importing Data

```{r}
exam_data <- read_csv("data/Exam_data.csv")
```

**Dataset Overview:**

```{r}
glimpse(exam_data)
```

```{r}
summary(exam_data)
```

## Initial Data Exploration

```{r}
# Calculate key statistics
n_students <- nrow(exam_data)
n_variables <- ncol(exam_data)
n_complete <- sum(complete.cases(exam_data))

# Subject statistics
maths_mean <- mean(exam_data$MATHS)
english_mean <- mean(exam_data$ENGLISH)
science_mean <- mean(exam_data$SCIENCE)

maths_sd <- sd(exam_data$MATHS)
english_sd <- sd(exam_data$ENGLISH)
science_sd <- sd(exam_data$SCIENCE)

# Categorical distributions
n_male <- sum(exam_data$GENDER == "Male")
n_female <- sum(exam_data$GENDER == "Female")
n_classes <- n_distinct(exam_data$CLASS)
n_races <- n_distinct(exam_data$RACE)

# Check for missing values
missing_count <- sum(is.na(exam_data))
```

::: {.callout-important}
## Dataset Characteristics

**Sample Size:** `r n_students` Primary 3 students from a local school

**Data Quality:** `r ifelse(missing_count == 0, "✓ Complete dataset with no missing values", paste("⚠", missing_count, "missing values detected"))`

**Variables:** `r n_variables` attributes
- **Categorical (4):** ID, CLASS, GENDER, RACE
- **Continuous (3):** MATHS, ENGLISH, SCIENCE

**Demographics:**
- Gender: `r n_male` males (`r round(n_male/n_students*100, 1)`%), `r n_female` females (`r round(n_female/n_students*100, 1)`%)
- Classes: `r n_classes` different classes
- Racial diversity: `r n_races` groups

**Academic Performance Overview:**
- Mathematics: μ = `r round(maths_mean, 1)`, σ = `r round(maths_sd, 1)`
- English: μ = `r round(english_mean, 1)`, σ = `r round(english_sd, 1)`
- Science: μ = `r round(science_mean, 1)`, σ = `r round(science_sd, 1)`

**Key Observation:** `r ifelse(max(c(maths_mean, english_mean, science_mean)) - min(c(maths_mean, english_mean, science_mean)) < 5, "Students show consistent performance across subjects", paste("There is a", round(max(c(maths_mean, english_mean, science_mean)) - min(c(maths_mean, english_mean, science_mean)), 1), "point difference between highest and lowest performing subjects"))`
:::

# Introducing ggplot2

[ggplot2](https://ggplot2.tidyverse.org/) is an R package for **declaratively** creating **data-driven** graphics based on ***The Grammar of Graphics***. It is part of the tidyverse family specially designed for visual exploration and communication.

## R Graphics vs ggplot2 Comparison

::: panel-tabset
## R Graphics

```{r}
hist(exam_data$MATHS)
```

## ggplot2

```{r}
ggplot(data=exam_data, aes(x = MATHS)) +
  geom_histogram(bins=10, 
                 boundary = 100,
                 color="black", 
                 fill="grey") +
  ggtitle("Distribution of Maths scores")
```
:::

```{r}
# Calculate distribution characteristics
maths_median <- median(exam_data$MATHS)
maths_mode <- as.numeric(names(sort(table(exam_data$MATHS), decreasing = TRUE)[1]))
skewness <- (maths_mean - maths_median) / maths_sd
```

::: {.callout-tip}
## Comparative Analysis: R Graphics vs ggplot2

**Distribution Shape:** The Mathematics scores show `r ifelse(abs(skewness) < 0.2, "a symmetric distribution", ifelse(skewness > 0, "positive skewness (right tail)", "negative skewness (left tail)"))` with:
- Mean: `r round(maths_mean, 1)`
- Median: `r round(maths_median, 1)`
- Skewness coefficient: `r round(skewness, 3)`

**ggplot2 Advantages Demonstrated:**
1. **Customization:** Precise control over bins (10), boundaries (100), and colors
2. **Clarity:** Clean axes with proper labels and titles
3. **Consistency:** Reproducible styling across all visualizations
4. **Extensibility:** Easy to add layers (we'll see this later)

The `r abs(round(maths_mean - maths_median, 1))` point difference between mean and median `r ifelse(abs(maths_mean - maths_median) < 2, "suggests minimal outlier influence", "indicates some asymmetry in the distribution")`.
:::

# Grammar of Graphics Principles

Grammar of Graphics defines rules for structuring mathematical and aesthetic elements into meaningful graphs. There are **seven grammatical elements** in ggplot2:

1. **Data**: The dataset being plotted
2. **Aesthetics**: Visual characteristics (position, colors, size, shape, transparency)
3. **Geometrics**: Visual elements (point, bar, line)
4. **Facets**: Multiple subsets of the same graph
5. **Statistics**: Statistical transformations (mean, confidence intervals)
6. **Coordinate systems**: The plane on which data are mapped
7. **Themes**: Non-data components (titles, legends, backgrounds)

# Essential Grammatical Elements: Data

```{r}
ggplot(data=exam_data)
```

::: {.callout-note}
## Understanding the Data Layer

**What happens here:**
- `ggplot()` initializes a ggplot object with `r n_students` observations
- Creates a blank canvas ready for visualization
- The dataset (`r n_variables` variables) is loaded but not yet mapped to visual elements
- This is the **foundation layer** upon which all other elements build

**Next step needed:** Define how variables map to visual properties (aesthetics)
:::

# Essential Grammatical Elements: Aesthetics

```{r}
ggplot(data=exam_data, aes(x= MATHS))
```

```{r}
# Calculate range for interpretation
maths_range <- max(exam_data$MATHS) - min(exam_data$MATHS)
maths_min <- min(exam_data$MATHS)
maths_max <- max(exam_data$MATHS)
```

::: {.callout-note}
## Understanding Aesthetic Mappings

**What changed:**
- Added `aes(x = MATHS)` mapping
- X-axis now displays Mathematics scores
- Scale ranges from `r maths_min` to `r maths_max` (range: `r maths_range` points)
- Axis label automatically generated from variable name

**Still missing:** Visual representation (geometric objects) to display the actual data

**Performance spread:** The `r maths_range`-point range from lowest to highest score indicates `r ifelse(maths_range > 80, "high performance variability", ifelse(maths_range > 60, "moderate performance variability", "relatively homogeneous performance"))` among students.
:::

# Essential Grammatical Elements: Geometric Objects

## geom_bar: Categorical Data Visualization

```{r}
ggplot(data=exam_data, aes(x=RACE)) +
  geom_bar()
```

```{r}
# Calculate racial distribution
race_counts <- exam_data %>%
  count(RACE) %>%
  arrange(desc(n))

largest_group <- race_counts$RACE[1]
largest_count <- race_counts$n[1]
smallest_group <- race_counts$RACE[nrow(race_counts)]
smallest_count <- race_counts$n[nrow(race_counts)]
diversity_ratio <- largest_count / smallest_count
```

::: {.callout-important}
## Racial Distribution Analysis

**Sample Composition:**
```{r}
#| echo: false
for (i in 1:nrow(race_counts)) {
  pct <- (race_counts$n[i] / n_students) * 100
  cat(paste0("- ", race_counts$RACE[i], ": ", race_counts$n[i], 
             " students (", round(pct, 1), "%)\n"))
}
```

**Diversity Assessment:**
- **Largest group:** `r largest_group` with `r largest_count` students (`r round(largest_count/n_students*100, 1)`%)
- **Smallest group:** `r smallest_group` with `r smallest_count` students (`r round(smallest_count/n_students*100, 1)`%)
- **Representation ratio:** `r round(diversity_ratio, 1)`:1 (largest to smallest)

**Interpretation:** The sample shows `r ifelse(diversity_ratio < 2, "relatively balanced racial representation", ifelse(diversity_ratio < 4, "moderate diversity with some groups larger than others", "significant imbalance with one group dominating"))`.
:::

## geom_dotplot: Fine-Grained Distribution

```{r}
ggplot(data=exam_data, aes(x = MATHS)) +
  geom_dotplot(dotsize = 0.5)
```

::: {.callout-warning}
## Misleading Y-Axis Issue

**Problem identified:** The y-axis shows "count" but represents stacking height, not frequency. With `r n_students` students, the y-axis scaling is **misleading and should be removed**.

**Solution:** Use `scale_y_continuous(NULL, breaks = NULL)` to eliminate the confusing scale.
:::

```{r}
ggplot(data=exam_data, aes(x = MATHS)) +
  geom_dotplot(binwidth=2.5, dotsize = 0.5) +
  scale_y_continuous(NULL, breaks = NULL)
```

```{r}
# Analyze clustering
score_bins <- cut(exam_data$MATHS, breaks = seq(0, 100, by = 10))
mode_bin <- names(sort(table(score_bins), decreasing = TRUE)[1])
```

::: {.callout-tip}
## Dot Plot Insights

**Visual Patterns Observed:**
- Each dot represents one student's Mathematics score
- Binwidth = 2.5 points reveals fine-grained clustering
- Highest concentration in the `r mode_bin` range
- Clear visualization of individual data points (n = `r n_students`)

**Advantage over histogram:** While histograms aggregate data, dot plots show **every individual observation**, making outliers and gaps immediately visible.

**Pattern:** `r ifelse(maths_sd < 15, "Dots cluster tightly around the mean, indicating consistent performance", "Dots spread widely, showing diverse performance levels")`
:::

## geom_histogram: Distribution Analysis

```{r}
ggplot(data=exam_data, aes(x = MATHS)) +
  geom_histogram()
```

::: {.callout-note}
**Default behavior:** 30 bins automatically chosen. With `r n_students` students and score range of `r maths_range`, each bin width ≈ `r round(maths_range/30, 2)` points.
:::

### Optimizing Histogram Bins

```{r}
ggplot(data=exam_data, aes(x= MATHS)) +
  geom_histogram(bins=20, color="black", fill="light blue")
```

```{r}
# Calculate optimal bin width using Sturges' rule
optimal_bins_sturges <- ceiling(log2(n_students) + 1)
# Freedman-Diaconis rule
iqr_maths <- IQR(exam_data$MATHS)
bin_width_fd <- 2 * iqr_maths / (n_students^(1/3))
optimal_bins_fd <- ceiling(maths_range / bin_width_fd)
```

::: {.callout-tip}
## Bin Selection Analysis

**Current choice:** 20 bins

**Statistical recommendations:**
- **Sturges' rule:** `r optimal_bins_sturges` bins (for n = `r n_students`)
- **Freedman-Diaconis rule:** `r optimal_bins_fd` bins (based on IQR = `r round(iqr_maths, 1)`)

**With 20 bins:** Each bin width = `r round(maths_range/20, 1)` points, providing `r ifelse(20 >= optimal_bins_sturges - 2 & 20 <= optimal_bins_sturges + 2, "appropriate granularity", ifelse(20 > optimal_bins_sturges + 2, "finer detail than statistically recommended", "coarser binning than optimal"))` for this dataset.

**Visual interpretation:** The histogram reveals `r ifelse(abs(skewness) < 0.2, "a roughly normal distribution", ifelse(skewness > 0, "slight right skew with more low performers", "slight left skew with more high performers"))` centered around `r round(maths_mean, 1)` points.
:::

### Aesthetic Mapping by Group

```{r}
ggplot(data=exam_data, aes(x= MATHS, fill = GENDER)) +
  geom_histogram(bins=20, color="grey30")
```

```{r}
# Gender-specific statistics
gender_stats <- exam_data %>%
  group_by(GENDER) %>%
  summarise(
    n = n(),
    mean_maths = mean(MATHS),
    sd_maths = sd(MATHS),
    median_maths = median(MATHS)
  )

male_mean <- gender_stats$mean_maths[gender_stats$GENDER == "Male"]
female_mean <- gender_stats$mean_maths[gender_stats$GENDER == "Female"]
gender_diff <- abs(male_mean - female_mean)
```

::: {.callout-important}
## Gender-Based Performance Comparison

**Sample distribution:**
- Males: `r gender_stats$n[gender_stats$GENDER == "Male"]` students
- Females: `r gender_stats$n[gender_stats$GENDER == "Female"]` students

**Mathematics performance:**
- **Male mean:** `r round(male_mean, 1)` (SD: `r round(gender_stats$sd_maths[gender_stats$GENDER == "Male"], 1)`)
- **Female mean:** `r round(female_mean, 1)` (SD: `r round(gender_stats$sd_maths[gender_stats$GENDER == "Female"], 1)`)
- **Difference:** `r round(gender_diff, 1)` points

**Interpretation:** The overlapping distributions show that `r ifelse(gender_diff < 3, "males and females perform similarly in Mathematics, with negligible difference", ifelse(male_mean > female_mean, paste0("males slightly outperform females by ", round(gender_diff, 1), " points on average"), paste0("females slightly outperform males by ", round(gender_diff, 1), " points on average")))`

**Visual insight:** The `r ifelse(gender_diff < 5, "extensive overlap", "moderate separation")` in the histogram confirms `r ifelse(gender_diff < 5, "gender is NOT a strong predictor of Mathematics performance", "some gender-based performance differences exist")`.
:::

## geom_density: Smooth Distribution Curves

```{r}
ggplot(data=exam_data, aes(x = MATHS)) +
  geom_density()
```

```{r}
# Calculate density peak
density_obj <- density(exam_data$MATHS)
peak_value <- density_obj$x[which.max(density_obj$y)]
```

::: {.callout-note}
## Kernel Density Estimation Analysis

**Method:** Smoothed version of histogram using kernel density estimation

**Peak location:** Approximately `r round(peak_value, 1)` points (mode)

**Compared to mean:** The peak at `r round(peak_value, 1)` vs mean of `r round(maths_mean, 1)` indicates `r ifelse(abs(peak_value - maths_mean) < 3, "symmetric distribution", ifelse(peak_value < maths_mean, "positive skew (tail extends right)", "negative skew (tail extends left)"))`

**Advantage:** Reveals underlying distribution pattern without arbitrary binning decisions that affect histograms.
:::

### Gender Comparison with Density Curves

```{r}
ggplot(data=exam_data, aes(x = MATHS, colour = GENDER)) +
  geom_density()
```

```{r}
# Calculate overlap coefficient
male_scores <- exam_data$MATHS[exam_data$GENDER == "Male"]
female_scores <- exam_data$MATHS[exam_data$GENDER == "Female"]

# Perform two-sample t-test
t_test_result <- t.test(male_scores, female_scores)
p_value <- t_test_result$p.value
```

::: {.callout-important}
## Statistical Gender Comparison

**Visual observation:** The density curves show `r ifelse(gender_diff < 3, "substantial overlap", ifelse(gender_diff < 7, "moderate overlap", "notable separation"))` between male and female distributions.

**Statistical test (Two-sample t-test):**
- **Mean difference:** `r round(abs(male_mean - female_mean), 2)` points
- **t-statistic:** `r round(t_test_result$statistic, 3)`
- **p-value:** `r round(p_value, 4)`

**Conclusion:** `r ifelse(p_value < 0.05, paste0("There IS a statistically significant difference (p = ", round(p_value, 4), " < 0.05) between male and female Mathematics performance. ", ifelse(male_mean > female_mean, "Males score significantly higher.", "Females score significantly higher.")), paste0("There is NO statistically significant difference (p = ", round(p_value, 4), " > 0.05) between genders. The ", round(gender_diff, 1), "-point difference can be attributed to random variation."))`

**Practical significance:** Even if statistically significant, a difference of `r round(gender_diff, 1)` points represents only `r round(gender_diff/maths_mean * 100, 1)`% of the mean score, indicating `r ifelse(gender_diff/maths_mean < 0.1, "minimal practical impact", "moderate practical relevance")`.
:::

## geom_boxplot: Five-Number Summary

```{r}
ggplot(data=exam_data, aes(y = MATHS, x= GENDER)) +
  geom_boxplot()
```

```{r}
# Calculate quartiles and outliers
quartiles_male <- quantile(male_scores, c(0.25, 0.5, 0.75))
quartiles_female <- quantile(female_scores, c(0.25, 0.5, 0.75))

iqr_male <- IQR(male_scores)
iqr_female <- IQR(female_scores)

# Detect outliers
outliers_male <- male_scores[male_scores < (quartiles_male[1] - 1.5*iqr_male) | 
                              male_scores > (quartiles_male[3] + 1.5*iqr_male)]
outliers_female <- female_scores[female_scores < (quartiles_female[1] - 1.5*iqr_female) | 
                                  female_scores > (quartiles_female[3] + 1.5*iqr_female)]
```

::: {.callout-tip}
## Box Plot Statistical Breakdown

**Male students (n = `r length(male_scores)`):**
- Q1 (25th percentile): `r round(quartiles_male[1], 1)`
- Median (50th percentile): `r round(quartiles_male[2], 1)`
- Q3 (75th percentile): `r round(quartiles_male[3], 1)`
- IQR (spread): `r round(iqr_male, 1)` points
- Outliers: `r length(outliers_male)` student(s)

**Female students (n = `r length(female_scores)`):**
- Q1: `r round(quartiles_female[1], 1)`
- Median: `r round(quartiles_female[2], 1)`
- Q3: `r round(quartiles_female[3], 1)`
- IQR: `r round(iqr_female, 1)` points
- Outliers: `r length(outliers_female)` student(s)

**Comparative insights:**
- **Median difference:** `r abs(round(quartiles_male[2] - quartiles_female[2], 1))` points
- **Spread comparison:** `r ifelse(iqr_male > iqr_female, paste0("Males show ", round((iqr_male/iqr_female - 1)*100, 1), "% more variability"), paste0("Females show ", round((iqr_female/iqr_male - 1)*100, 1), "% more variability"))`
- **Middle 50%:** Males range `r round(iqr_male, 1)` points vs Females `r round(iqr_female, 1)` points

**Outlier analysis:** `r ifelse(length(outliers_male) + length(outliers_female) == 0, "No outliers detected - scores follow normal pattern", paste0(length(outliers_male) + length(outliers_female), " outlier(s) identified, representing ", round((length(outliers_male) + length(outliers_female))/n_students*100, 1), "% of students"))`
:::

### Notched Box Plots

```{r}
ggplot(data=exam_data, aes(y = MATHS, x= GENDER)) +
  geom_boxplot(notch=TRUE)
```

::: {.callout-note}
## Notched Box Plot Interpretation

**Purpose of notches:** Provide visual assessment of whether medians differ significantly.

**Rule:** If notches **do not overlap**, there is strong evidence (95% confidence) that medians differ.

**Observation:** The notches `r ifelse(abs(quartiles_male[2] - quartiles_female[2]) > 1.58 * sqrt((iqr_male^2/length(male_scores) + iqr_female^2/length(female_scores))), "DO NOT overlap", "DO overlap")`, suggesting medians are `r ifelse(abs(quartiles_male[2] - quartiles_female[2]) > 1.58 * sqrt((iqr_male^2/length(male_scores) + iqr_female^2/length(female_scores))), "statistically different", "not statistically different")`.

**Alignment with t-test:** This visual test `r ifelse((p_value < 0.05) == (abs(quartiles_male[2] - quartiles_female[2]) > 1.58 * sqrt((iqr_male^2/length(male_scores) + iqr_female^2/length(female_scores)))), "CONFIRMS", "contradicts")` our earlier t-test conclusion (p = `r round(p_value, 4)`).
:::

## geom_violin: Distribution Shape Comparison

```{r}
ggplot(data=exam_data, aes(y = MATHS, x= GENDER)) +
  geom_violin()
```

::: {.callout-tip}
## Violin Plot Advantages

**What violin plots reveal:**
1. **Full distribution shape:** Unlike box plots (5-number summary only), violin plots show entire density
2. **Multi-modal patterns:** Can detect multiple peaks in data
3. **Asymmetry:** Clear visualization of skewness

**Current observations:**
- `r ifelse(iqr_male > iqr_female, "Male distribution appears wider", "Female distribution appears wider")`
- Both show `r ifelse(abs(skewness) < 0.3, "roughly symmetric shapes", ifelse(skewness > 0, "slight bottom-heavy shapes", "slight top-heavy shapes"))`
- No obvious bi-modal patterns (would appear as two bulges)

**Comparison value:** With `r n_students` students split by gender, violin plots provide richer detail than box plots for understanding performance distribution differences.
:::

## geom_point: Scatter Plots for Relationships

```{r}
ggplot(data=exam_data, aes(x= MATHS, y=ENGLISH)) +
  geom_point()
```

```{r}
# Calculate correlation
cor_maths_english <- cor(exam_data$MATHS, exam_data$ENGLISH)
cor_test <- cor.test(exam_data$MATHS, exam_data$ENGLISH)

# Count high achievers in both subjects
both_high <- sum(exam_data$MATHS >= 75 & exam_data$ENGLISH >= 75)
both_low <- sum(exam_data$MATHS < 50 & exam_data$ENGLISH < 50)
```

::: {.callout-important}
## Subject Correlation Analysis

**Pearson correlation coefficient:** r = `r round(cor_maths_english, 3)`

**Interpretation:** `r ifelse(abs(cor_maths_english) > 0.7, "Strong positive correlation", ifelse(abs(cor_maths_english) > 0.4, "Moderate positive correlation", "Weak correlation"))` - students who score high in Mathematics tend to score high in English.

**Statistical significance:**
- p-value: `r round(cor_test$p.value, 6)`
- `r ifelse(cor_test$p.value < 0.001, "Highly significant (p < 0.001)", ifelse(cor_test$p.value < 0.05, "Statistically significant (p < 0.05)", "Not statistically significant"))`

**Shared variance:** r² = `r round(cor_maths_english^2, 3)`, meaning **`r round(cor_maths_english^2 * 100, 1)`%** of variability in English scores can be explained by Mathematics performance.

**Pattern analysis:**
- **High achievers (≥75 in both):** `r both_high` students (`r round(both_high/n_students*100, 1)`%)
- **Low achievers (<50 in both):** `r both_low` students (`r round(both_low/n_students*100, 1)`%)
- **Consistency:** `r round((both_high + both_low)/n_students*100, 1)`% of students show consistent performance across subjects

**Educational implication:** The r = `r round(cor_maths_english, 3)` correlation suggests `r ifelse(cor_maths_english > 0.7, "strong underlying academic abilities that transfer across subjects", ifelse(cor_maths_english > 0.4, "moderate skill transfer between subjects", "relatively independent subject-specific abilities"))`.
:::

## Combining Geometric Objects

```{r}
ggplot(data=exam_data, aes(y = MATHS, x= GENDER)) +
  geom_boxplot() +
  geom_point(position="jitter", size = 0.5)
```

```{r}
# Count points by gender for jitter analysis
n_male_points <- sum(exam_data$GENDER == "Male")
n_female_points <- sum(exam_data$GENDER == "Female")
```

::: {.callout-tip}
## Layered Visualization Benefits

**Combined elements:**
1. **Box plot layer:** Shows median, quartiles, outliers (`r length(outliers_male) + length(outliers_female)` outliers)
2. **Jittered points layer:** Shows all `r n_students` individual observations

**Jittering rationale:**
- Without jitter: Points stack vertically (overplotting)
- With jitter: Horizontal randomization reveals density
- Point size = 0.5 to reduce overlap

**What this reveals:**
- **Male group:** `r n_male_points` points show `r ifelse(iqr_male > 20, "wide spread", "concentrated distribution")`
- **Female group:** `r n_female_points` points show `r ifelse(iqr_female > 20, "wide spread", "concentrated distribution")`
- Individual points confirm the box plot's summary statistics

**Advantage:** Combines **summary statistics** (box plot) with **raw data** (points) for complete picture.
:::

# Essential Grammatical Elements: Statistics

## Adding Statistical Summaries

```{r}
ggplot(data=exam_data, aes(y = MATHS, x= GENDER)) +
  geom_boxplot()
```

The boxplot above shows medians but not means. Let's add them:

```{r}
ggplot(data=exam_data, aes(y = MATHS, x= GENDER)) +
  geom_boxplot() +
  stat_summary(geom = "point", fun.y="mean", colour ="red", size=4)
```

```{r}
# Calculate mean positions
mean_diff_from_median_male <- male_mean - quartiles_male[2]
mean_diff_from_median_female <- female_mean - quartiles_female[2]
```

::: {.callout-note}
## Mean vs Median Analysis

**Red points show means overlaid on box plots (which show medians)**

**Male students:**
- Mean: `r round(male_mean, 1)` (red point)
- Median: `r round(quartiles_male[2], 1)` (box line)
- Difference: `r round(abs(mean_diff_from_median_male), 1)` points

**Female students:**
- Mean: `r round(female_mean, 1)` (red point)
- Median: `r round(quartiles_female[2], 1)` (box line)
- Difference: `r round(abs(mean_diff_from_median_female), 1)` points

**Interpretation:** `r ifelse(max(abs(mean_diff_from_median_male), abs(mean_diff_from_median_female)) < 2, "Means and medians are nearly identical, indicating symmetric distributions without outlier influence", "Mean-median differences suggest some skewness or outlier impact")`

**Why this matters:** When mean ≈ median, either measure adequately represents "typical" performance. When they differ substantially, median is more robust to outliers.
:::

## Alternative stat() Approach

```{r}
ggplot(data=exam_data, aes(y = MATHS, x= GENDER)) +
  geom_boxplot() +
  geom_point(stat="summary", fun.y="mean", colour ="red", size=4)
```

::: {.callout-note}
**Two approaches, same result:**
- `stat_summary(geom = "point", ...)` - stat function, specify geom
- `geom_point(stat = "summary", ...)` - geom function, specify stat

Both achieve identical output, demonstrating ggplot2's flexible layering system.
:::

## Adding Regression Lines

```{r}
ggplot(data=exam_data, aes(x= MATHS, y=ENGLISH)) +
  geom_point() +
  geom_smooth(size=0.5)
```

```{r}
# Calculate smoothing window
n_effective <- n_students * 0.75  # LOESS default span
```

::: {.callout-note}
## LOESS Smoothing Analysis

**Method:** Local polynomial regression (LOESS - LOcally Estimated Scatterplot Smoothing)

**How it works:** Fits polynomial curves to localized subsets of data

**Confidence band (gray area):** 95% confidence interval for the smooth curve

**Observation:** The smooth curve is `r ifelse(cor_maths_english > 0.8, "nearly linear", "curved")`, suggesting `r ifelse(cor_maths_english > 0.8, "a linear relationship is appropriate", "non-linear components exist")`

**Effective sample size:** LOESS uses ≈`r round(n_effective)` nearest neighbors for each point
:::

### Linear Regression Alternative

```{r}
ggplot(data=exam_data, aes(x= MATHS, y=ENGLISH)) +
  geom_point() +
  geom_smooth(method=lm, size=0.5)
```

```{r}
# Fit linear model
lm_model <- lm(ENGLISH ~ MATHS, data = exam_data)
slope <- coef(lm_model)[2]
intercept <- coef(lm_model)[1]
r_squared <- summary(lm_model)$r.squared
residual_se <- summary(lm_model)$sigma
```

::: {.callout-important}
## Linear Regression Analysis

**Model:** ENGLISH = `r round(intercept, 2)` + `r round(slope, 3)` × MATHS

**Interpretation:**
- **Intercept:** `r round(intercept, 2)` - predicted English score when Maths = 0 (extrapolation, not meaningful)
- **Slope:** `r round(slope, 3)` - for every 1-point increase in Maths, English increases by `r round(slope, 3)` points
- **R-squared:** `r round(r_squared, 3)` - **`r round(r_squared*100, 1)`%** of English score variance explained by Maths

**Model fit quality:**
- Residual standard error: `r round(residual_se, 2)` points
- Typical prediction is off by ±`r round(residual_se, 2)` points

**Practical example:**
- Student scoring 70 in Maths → predicted English: `r round(intercept + slope*70, 1)` ± `r round(residual_se, 2)`
- Student scoring 90 in Maths → predicted English: `r round(intercept + slope*90, 1)` ± `r round(residual_se, 2)`

**Model adequacy:** R² = `r round(r_squared, 3)` indicates `r ifelse(r_squared > 0.7, "strong predictive power", ifelse(r_squared > 0.4, "moderate predictive power", "weak predictive power"))` - Maths score is `r ifelse(r_squared > 0.7, "an excellent", ifelse(r_squared > 0.4, "a useful", "a limited"))` predictor of English performance.
:::

# Essential Grammatical Elements: Facets

## facet_wrap: Trellis Plots

```{r}
ggplot(data=exam_data, aes(x= MATHS)) +
  geom_histogram(bins=20) +
  facet_wrap(~ CLASS)
```

```{r}
# Calculate class-level statistics
class_stats <- exam_data %>%
  group_by(CLASS) %>%
  summarise(
    n = n(),
    mean_maths = mean(MATHS),
    sd_maths = sd(MATHS),
    median_maths = median(MATHS)
  ) %>%
  arrange(desc(mean_maths))

best_class <- class_stats$CLASS[1]
worst_class <- class_stats$CLASS[nrow(class_stats)]
class_gap <- class_stats$mean_maths[1] - class_stats$mean_maths[nrow(class_stats)]
```

::: {.callout-important}
## Class-Level Performance Analysis

**Number of classes:** `r n_classes`

**Class performance ranking (by mean Maths score):**
```{r}
#| echo: false
for (i in 1:nrow(class_stats)) {
  cat(paste0(i, ". ", class_stats$CLASS[i], ": ",
             "μ = ", round(class_stats$mean_maths[i], 1),
             ", σ = ", round(class_stats$sd_maths[i], 1),
             ", n = ", class_stats$n[i], "\n"))
}
```

**Key findings:**
- **Highest performing:** `r best_class` (μ = `r round(class_stats$mean_maths[1], 1)`)
- **Lowest performing:** `r worst_class` (μ = `r round(class_stats$mean_maths[nrow(class_stats)], 1)`)
- **Performance gap:** `r round(class_gap, 1)` points between top and bottom classes

**Interpretation:** The `r round(class_gap, 1)`-point gap represents `r round(class_gap/maths_mean * 100, 1)`% of overall mean, indicating `r ifelse(class_gap < 10, "relatively consistent teaching quality across classes", ifelse(class_gap < 20, "moderate variation in class performance", "significant inter-class disparities requiring investigation"))`

**Distribution shapes:** Visual inspection shows `r ifelse(sd(class_stats$mean_maths) < 5, "similar distribution patterns", "varying distribution patterns")` across classes, with `r sum(class_stats$mean_maths > maths_mean)` classes above overall mean.
:::

## facet_grid: Matrix Layout

```{r}
ggplot(data=exam_data, aes(x= MATHS)) +
  geom_histogram(bins=20) +
  facet_grid(~ CLASS)
```

::: {.callout-note}
## facet_wrap vs facet_grid

**facet_wrap (previous plot):**
- Wraps panels into 2D grid automatically
- Better for many categories (uses space efficiently)
- Our `r n_classes` classes wrapped into multiple rows

**facet_grid (current plot):**
- Forces single row/column
- Better for 1-2 categorical variables
- All `r n_classes` classes in one row (may be cramped)

**Recommendation:** With `r n_classes` classes, `facet_wrap()` provides better readability.
:::

# Essential Grammatical Elements: Coordinates

## Flipping Coordinates

```{r}
ggplot(data=exam_data, aes(x=RACE)) +
  geom_bar()
```

```{r}
ggplot(data=exam_data, aes(x=RACE)) +
  geom_bar() +
  coord_flip()
```

::: {.callout-tip}
## When to Flip Coordinates

**Vertical bars (default):**
- Good for few categories
- Category names fit horizontally

**Horizontal bars (coord_flip):**
- **Better for:** Long category names, many categories, easier comparison
- **This case:** `r n_races` racial groups - horizontal makes category labels more readable
- **Comparison:** Easier to compare bar lengths left-to-right than bottom-to-top

**Readability gain:** Flipping allows full category names without rotation or truncation.
:::

## Setting Axis Ranges

```{r}
ggplot(data=exam_data, aes(x= MATHS, y=ENGLISH)) +
  geom_point() +
  geom_smooth(method=lm, size=0.5)
```

```{r}
# Calculate current axis ranges
actual_x_range <- c(min(exam_data$MATHS), max(exam_data$MATHS))
actual_y_range <- c(min(exam_data$ENGLISH), max(exam_data$ENGLISH))
```

::: {.callout-warning}
## Misleading Aspect Ratio Issue

**Problem identified:**
- X-axis range: `r round(actual_x_range[1], 1)` to `r round(actual_x_range[2], 1)` (range: `r round(actual_x_range[2] - actual_x_range[1], 1)`)
- Y-axis range: `r round(actual_y_range[1], 1)` to `r round(actual_y_range[2], 1)` (range: `r round(actual_y_range[2] - actual_y_range[1], 1)`)

**Why this misleads:** Different axis ranges distort the visual slope of the regression line. A 1-point change in Maths appears different from a 1-point change in English.

**Solution:** Use equal scales (0-100 for both) to accurately represent relationship.
:::

```{r}
ggplot(data=exam_data, aes(x= MATHS, y=ENGLISH)) +
  geom_point() +
  geom_smooth(method=lm, size=0.5) +
  coord_cartesian(xlim=c(0,100), ylim=c(0,100))
```

::: {.callout-tip}
## Fixed Coordinates Analysis

**Improvement:** Both axes now 0-100

**Visual accuracy:** The regression line's angle now truly represents the relationship strength

**Slope interpretation:** With equal scales, slope = `r round(slope, 3)` means:
- `r ifelse(slope < 0.9, "English gains are slightly less than Maths gains", ifelse(slope > 1.1, "English gains are slightly more than Maths gains", "Nearly 1:1 relationship - gains transfer equally"))`

**Honest visualization:** Equal scales prevent exaggerating or minimizing the correlation (r = `r round(cor_maths_english, 3)`)
:::

# Essential Grammatical Elements: Themes

## Built-in Theme Comparison

```{r}
ggplot(data=exam_data, aes(x=RACE)) +
  geom_bar() +
  coord_flip() +
  theme_gray()
```

```{r}
ggplot(data=exam_data, aes(x=RACE)) +
  geom_bar() +
  coord_flip() +
  theme_classic()
```

```{r}
ggplot(data=exam_data, aes(x=RACE)) +
  geom_bar() +
  coord_flip() +
  theme_minimal()
```

::: {.callout-tip}
## Theme Selection Guidelines

**theme_gray() (default):**
- Gray background with white gridlines
- Best for: Screen viewing, presentations
- Drawback: Less professional for publications

**theme_classic():**
- White background, no gridlines, only axes
- Best for: Academic papers, traditional plots
- Advantage: Minimal ink, maximum data focus

**theme_minimal():**
- White background with subtle gridlines
- Best for: Modern reports, dashboards
- Balance: Clean but retains helpful gridlines

**For this data (n = `r n_students`, `r n_races` racial groups):**
All three effectively display the distribution. Choice depends on context:
- **Publication:** theme_classic()
- **Presentation:** theme_gray()
- **Web/dashboard:** theme_minimal()
:::

# Comprehensive Summary Analysis

```{r}
# Calculate comprehensive statistics
overall_performance <- data.frame(
  Subject = c("Mathematics", "English", "Science"),
  Mean = c(maths_mean, english_mean, science_mean),
  SD = c(maths_sd, english_sd, science_sd),
  CV = c(maths_sd/maths_mean, english_sd/english_mean, science_sd/science_mean) * 100
)

# Correlations
cor_matrix <- cor(exam_data[, c("MATHS", "ENGLISH", "SCIENCE")])
cor_me <- cor_matrix["MATHS", "ENGLISH"]
cor_ms <- cor_matrix["MATHS", "SCIENCE"]
cor_es <- cor_matrix["ENGLISH", "SCIENCE"]
avg_correlation <- mean(c(cor_me, cor_ms, cor_es))
```

::: {.callout-important}
## Final Data-Driven Insights

### Overall Academic Performance

**Subject comparison:**
```{r}
#| echo: false
for (i in 1:nrow(overall_performance)) {
  cat(paste0("- **", overall_performance$Subject[i], ":** ",
             "μ = ", round(overall_performance$Mean[i], 1),
             ", σ = ", round(overall_performance$SD[i], 1),
             ", CV = ", round(overall_performance$CV[i], 1), "%\n"))
}
```

**Best performing subject:** `r overall_performance$Subject[which.max(overall_performance$Mean)]`

**Most consistent performance:** `r overall_performance$Subject[which.min(overall_performance$CV)]` (lowest CV)

### Inter-Subject Relationships

**Correlation matrix:**
- Maths ↔ English: r = `r round(cor_me, 3)`
- Maths ↔ Science: r = `r round(cor_ms, 3)`
- English ↔ Science: r = `r round(cor_es, 3)`

**Average correlation:** `r round(avg_correlation, 3)`

**Interpretation:** `r ifelse(avg_correlation > 0.7, "Strong positive correlations across all subjects indicate general academic ability dominates subject-specific skills", ifelse(avg_correlation > 0.4, "Moderate correlations suggest both general ability and subject-specific factors matter", "Weak correlations indicate relatively independent subject-specific abilities"))`

### Demographic Factors

**Gender impact:**
- Mathematics difference: `r round(gender_diff, 1)` points (`r ifelse(p_value < 0.05, "statistically significant", "not statistically significant")`)
- Practical significance: `r ifelse(gender_diff < 5, "Minimal - gender not a strong predictor", ifelse(gender_diff < 10, "Moderate - some gender influence", "Substantial - gender matters"))`

**Class-level variation:**
- Inter-class gap: `r round(class_gap, 1)` points
- Coefficient of variation: `r round(sd(class_stats$mean_maths)/mean(class_stats$mean_maths)*100, 1)`%
- Assessment: `r ifelse(class_gap < 15, "Acceptable consistency across classes", ifelse(class_gap < 30, "Moderate variation - monitor teaching quality", "High variation - immediate intervention needed"))`

### Key Takeaways

1. **Academic consistency:** Average correlation of `r round(avg_correlation, 3)` means students performing well in one subject typically excel in others

2. **Performance spread:** Standard deviations of `r round(mean(overall_performance$SD), 1)` points (average) indicate `r ifelse(mean(overall_performance$SD) < 15, "relatively homogeneous class", ifelse(mean(overall_performance$SD) < 20, "moderate diversity in abilities", "high diversity requiring differentiated instruction"))`

3. **Demographic equity:** `r ifelse(max(gender_diff, class_gap) < 10, "No major demographic disparities detected", paste0("Attention needed: ", round(max(gender_diff, class_gap), 1), "-point gap in ", ifelse(gender_diff > class_gap, "gender", "class performance")))`

4. **Data quality:** `r n_students` complete observations with no missing data provide robust foundation for analysis
:::

# References

-   Hadley Wickham (2023) [**ggplot2: Elegant Graphics for Data Analysis**](https://ggplot2-book.org/). Online 3rd edition.
-   Winston Chang (2013) [**R Graphics Cookbook 2nd edition**](https://r-graphics.org/). Online version.
-   Healy, Kieran (2019) [**Data Visualization: A practical introduction**](https://socviz.co/). Online version
-   [Learning ggplot2 on Paper – Components](https://henrywang.nl/learning-ggplot2-on-paper-components/)
-   [Learning ggplot2 on Paper – Layer](https://henrywang.nl/learning-ggplot2-on-paper-layer/)
-   [Learning ggplot2 on Paper – Scale](https://henrywang.nl/tag/learning-ggplot2-on-paper/)




