---
title: "Hands-on Exercise 1"
subtitle: "A Layered Grammar of Graphics: ggplot2 methods"
author: "Thi Thanh Van Tran"
date: "April 16, 2026"
date-modified: "April 16, 2026"
format:
  html:
    toc: true
    number-sections: true
    number-depth: 4
    code-fold: true
    code-summary: "Show code"
    code-tools: false
    css: styles.css
    theme: cosmo
    page-layout: article
execute:
  eval: true
  echo: true
  warning: false
  freeze: false
knitr:
  opts_chunk:
    out.width: "90%"
    fig.align: "center"
    dpi: 120        
---

::: welcome-section
# Welcome to My ggplot2 Journey

In this hands-on exercise, I explored the **Grammar of Graphics** using ggplot2 to create meaningful visualizations. Through analyzing exam data from Primary 3 students, I discovered the power of layered graphics and data-driven insights.

**Let's visualize data beautifully**
:::

**Learning Objectives:**
-   Master basic principles and essential components of ggplot2
-   Plot statistical graphics based on the principle of Layered Grammar of Graphics
-   Conduct data-driven analysis with dynamic insights
-   Extract meaningful patterns from visualizations

# Getting Started

## Installing and Loading Libraries

```{r}
#| code-fold: false
pacman::p_load(tidyverse, patchwork, knitr, ggplot2, reshape2)
```

::: {.callout-note}
## Package Installation

The `pacman::p_load()` function automatically installs missing packages and loads them into R. This ensures reproducibility across different environments.
:::

## Importing the Dataset

```{r}
#| code-fold: false
exam_data <- read_csv("data/Exam_data.csv")
```

# Visual Data Exploration

::: {.panel-tabset}

## Dataset Overview

```{r}
#| code-fold: true
#| code-summary: "Show code"
#| fig-width: 14
#| fig-height: 9

n_students <- nrow(exam_data)
n_complete <- sum(complete.cases(exam_data))

# Variable types
var_types <- data.frame(
  Type = sapply(exam_data, function(x) class(x)[1])
) %>%
  count(Type)

p2 <- ggplot(var_types, aes(x = reorder(Type, -n), y = n, fill = Type)) +
  geom_col(alpha = 0.8, width = 0.6) +
  geom_text(aes(label = n), vjust = -0.5, fontface = "bold", size = 5) +
  scale_fill_manual(values = c("character" = "#fbcfe8", "numeric" = "#bfdbfe")) +
  labs(title = "Variable Types", x = "Type", y = "Count") +
  ylim(0, max(var_types$n) * 1.2) +
  theme_minimal() +
  theme(plot.title = element_text(hjust = 0.5, face = "bold", size = 14),
        legend.position = "none")

# Sample size
p3 <- ggplot(data.frame(x = 1, y = 1), aes(x, y)) +
  geom_tile(fill = "#fce7f3", alpha = 0.5, width = 0.9, height = 0.9) +
  geom_text(label = paste0("n = ", n_students, "\nstudents"), 
            size = 8, fontface = "bold", color = "#831843") +
  labs(title = "Sample Size") +
  theme_void() +
  theme(plot.title = element_text(hjust = 0.5, face = "bold", size = 14))

# Missing values heatmap
missing_df <- melt(is.na(exam_data))

p4 <- ggplot(missing_df, aes(x = Var2, y = Var1, fill = value)) +
  geom_tile(color = "white", size = 0.05) +
  scale_fill_manual(values = c("FALSE" = "#10b981", "TRUE" = "#ef4444"),
                    labels = c("Complete", "Missing"), name = "Status") +
  labs(title = "Missing Value Heatmap", x = "Variables", y = "Observations") +
  theme_minimal() +
  theme(plot.title = element_text(hjust = 0.5, face = "bold", size = 14),
        axis.text.y = element_blank(),
        axis.ticks.y = element_blank(),
        axis.text.x = element_text(angle = 45, hjust = 1),
        panel.grid = element_blank())

(p3 | p2) / p4 +
  plot_annotation(title = "Dataset Overview at a Glance",
                  theme = theme(plot.title = element_text(size = 16, face = "bold", hjust = 0.5))) +
  plot_layout(heights = c(1, 1.2))
```

> **Judgment:** The dataset contains `r n_students` students with a healthy mix of numeric and character variables. The missing value heatmap confirms largely complete records, making the data suitable for analysis with minimal imputation concerns.

## Data Insights

```{r}
#| code-fold: true
#| code-summary: "Show code"
#| fig-width: 14
#| fig-height: 10

# Pastel palette — soft, non-viridis
pastel_palette <- c(
  "#FFB3C1", "#FFD6A5", "#CAFFBF", "#A0C4FF",
  "#BDB2FF", "#FFC6FF", "#FDFFB6", "#B5EAD7",
  "#F2C6DE", "#C9E4CA"
)

categorical_vars <- setdiff(
  names(exam_data)[sapply(exam_data, function(x) 
    is.character(x) || is.factor(x))],
  "ID"
)

plot_list <- lapply(categorical_vars, function(var) {
  
  plot_data <- exam_data %>%
    count(.data[[var]]) %>%
    mutate(pct = n / sum(n) * 100)
  
  max_label_length <- max(nchar(as.character(plot_data[[var]])))
  n_cats <- nrow(plot_data)
  fill_colors <- rep_len(pastel_palette, n_cats)
  
  p <- ggplot(plot_data, aes(x = .data[[var]], y = n, fill = .data[[var]])) +
    geom_col(alpha = 0.9, width = 0.6, color = "white") +
    geom_text(aes(label = paste0(n, "\n(", round(pct, 1), "%)")),
              vjust = if(max_label_length > 8) -0.2 else -0.3,
              hjust = if(max_label_length > 8) 1.1 else 0.5,
              size = 3, fontface = "bold", color = "#5a2d3a") +
    scale_fill_manual(values = fill_colors) +
    labs(title = paste(var, "Distribution"), y = "Count", x = "") +
    ylim(0, max(plot_data$n) * 1.18) +
    theme_minimal(base_size = 10) +
    theme(
      plot.title    = element_text(hjust = 0.5, face = "bold", size = 10, color = "#7a2840"),
      legend.position = "none",
      axis.text     = element_text(size = 8, color = "#5a2d3a"),
      axis.title    = element_text(size = 8, color = "#5a2d3a"),
      panel.grid.minor = element_blank(),
      axis.text.x   = element_text(
        angle = if(max_label_length > 8) 45 else 0, 
        hjust = if(max_label_length > 8) 1 else 0.5
      )
    )
  
  if(max_label_length > 8) {
    p <- p + coord_flip() +
      theme(axis.text.x = element_text(angle = 0, hjust = 0.5))
  }
  
  return(p)
})

wrap_plots(plot_list, ncol = 2) +
  plot_annotation(
    title = "Categorical Variables Distribution",
    theme = theme(
      plot.title = element_text(size = 12, face = "bold", hjust = 0.5, color = "#7a2840")
    )
  )
```

::: {.callout-note}
## Distribution Summary

```{r}
#| echo: false
# Dynamic summary for all categorical variables
for(var in categorical_vars) {
  cat("**", var, ":**\n")
  counts <- table(exam_data[[var]])
  for(i in seq_along(counts)) {
    pct <- (counts[i] / sum(counts)) * 100
    cat("- ", names(counts)[i], ": ", counts[i], " (", round(pct, 1), "%)\n", sep = "")
  }
  cat("\n")
}
```

**Key Finding**: The dataset shows balanced distribution across most categorical variables, enabling robust comparative analysis without major sample size disparities.

**My Learning**: Visualizing demographics immediately reveals representation patterns. The balanced gender split is particularly valuable for unbiased gender analysis later.
:::

> **Judgment:** The categorical distributions are broadly balanced across groups, which strengthens the validity of any subgroup comparisons. No single category dominates disproportionately, reducing the risk of skewed conclusions driven by unequal representation.

## Data Summary

```{r}
#| code-fold: true
#| code-summary: "Show code"

# Quick glimpse
glimpse(exam_data)
```

```{r}
#| code-fold: true
#| code-summary: "Show code"

# Statistical summary for numeric variables
summary(exam_data %>% select(where(is.numeric)))
```

> **Judgment:** The summary statistics reveal the central tendency and spread of each numeric variable. Comparing means and medians helps flag potential skewness, while min/max values highlight any outliers that may warrant further investigation before modelling.

:::

:::

# Academic Performance Overview

```{r}
#| code-fold: true
#| code-summary: "Show code"
#| fig-width: 14
#| fig-height: 8

# Get numeric variables automatically
numeric_vars <- exam_data %>% select(where(is.numeric))

# Correlation matrix
cor_matrix <- cor(numeric_vars)
cor_data <- as.data.frame(as.table(cor_matrix))
names(cor_data) <- c("Var1", "Var2", "Correlation")

# Correlation heatmap
p_cor <- ggplot(cor_data, aes(x = Var1, y = Var2, fill = Correlation)) +
  geom_tile(color = "white", size = 1.5) +
  geom_text(aes(label = round(Correlation, 2)), size = 5, fontface = "bold") +
  scale_fill_gradient2(low = "#fce7f3", mid = "#f9a8d4", high = "#ec4899",
                       midpoint = 0.5, limits = c(0, 1)) +
  labs(title = "Inter-Subject Correlation Matrix", x = "", y = "") +
  theme_minimal() +
  theme(plot.title = element_text(hjust = 0.5, face = "bold", size = 14))

# Distribution comparison
perf_long <- numeric_vars %>%
  pivot_longer(everything(), names_to = "Subject", values_to = "Score")

p_dist <- ggplot(perf_long, aes(x = Score, fill = Subject)) +
  geom_density(alpha = 0.6, size = 1) +
  scale_fill_manual(values = c("MATHS" = "#bfdbfe", 
                                "ENGLISH" = "#bbf7d0", 
                                "SCIENCE" = "#e9d5ff")) +
  labs(title = "Score Distributions", x = "Score", y = "Density") +
  theme_minimal() +
  theme(plot.title = element_text(hjust = 0.5, face = "bold", size = 13),
        legend.position = "top")

p_box <- ggplot(perf_long, aes(x = Subject, y = Score, fill = Subject)) +
  geom_violin(alpha = 0.3) +
  geom_boxplot(width = 0.2, alpha = 0.8) +
  stat_summary(fun = mean, geom = "point", shape = 23, size = 4, fill = "#831843") +
  scale_fill_manual(values = c("MATHS" = "#bfdbfe", 
                                "ENGLISH" = "#bbf7d0", 
                                "SCIENCE" = "#e9d5ff")) +
  labs(title = "Distribution by Subject", y = "Score", x = "") +
  theme_minimal() +
  theme(plot.title = element_text(hjust = 0.5, face = "bold", size = 13),
        legend.position = "none")

(p_cor | p_dist) / p_box +
  plot_annotation(title = "Academic Performance Overview",
                  theme = theme(plot.title = element_text(size = 16, face = "bold", hjust = 0.5)))
```

```{r}
#| echo: false
avg_cor <- mean(cor_matrix[lower.tri(cor_matrix)])
```

::: {.callout-important}
## Key Insights

**Average correlation**: `r round(avg_cor, 3)` → Strong positive relationships

**Implication**: Students who excel in one subject tend to excel in others, suggesting general academic ability rather than subject-specific talent.
:::


## Academic Performance: Distribution Patterns

```{r}
#| code-fold: true
#| code-summary: "Show code"
#| fig-width: 12
#| fig-height: 8

# Calculate statistics
stats_summary <- exam_data %>%
  summarise(across(c(MATHS, ENGLISH, SCIENCE), 
                   list(mean = ~mean(., na.rm = TRUE),
                        sd = ~sd(., na.rm = TRUE),
                        median = ~median(., na.rm = TRUE))))

# Reshape for visualization
perf_data <- exam_data %>%
  select(MATHS, ENGLISH, SCIENCE) %>%
  pivot_longer(everything(), names_to = "Subject", values_to = "Score")

# Density plots
p1 <- ggplot(perf_data, aes(x = Score, fill = Subject)) +
  geom_density(alpha = 0.6, size = 1) +
  scale_fill_manual(values = c("MATHS" = "#bfdbfe", 
                                "ENGLISH" = "#bbf7d0", 
                                "SCIENCE" = "#e9d5ff")) +
  labs(title = "Score Distributions Across Subjects",
       x = "Score", y = "Density") +
  theme_minimal() +
  theme(plot.title = element_text(face = "bold", hjust = 0.5, size = 12),
        legend.position = "top")

# Box plots with violin
p2 <- ggplot(perf_data, aes(x = Subject, y = Score, fill = Subject)) +
  geom_violin(alpha = 0.4, trim = FALSE) +
  geom_boxplot(width = 0.2, alpha = 0.8, outlier.shape = 21, 
               outlier.fill = "#831843", outlier.size = 3) +
  stat_summary(fun = mean, geom = "point", shape = 23, 
               size = 4, fill = "#831843") +
  scale_fill_manual(values = c("MATHS" = "#bfdbfe", 
                                "ENGLISH" = "#bbf7d0", 
                                "SCIENCE" = "#e9d5ff")) +
  labs(title = "Distribution Shape & Statistics",
       y = "Score") +
  theme_minimal() +
  theme(legend.position = "none",
        plot.title = element_text(face = "bold", hjust = 0.5, size = 12))

# Statistical summary visual
stats_long <- data.frame(
  Subject = rep(c("MATHS", "ENGLISH", "SCIENCE"), 2),
  Metric = rep(c("Mean", "SD"), each = 3),
  Value = c(stats_summary$MATHS_mean, stats_summary$ENGLISH_mean, stats_summary$SCIENCE_mean,
            stats_summary$MATHS_sd, stats_summary$ENGLISH_sd, stats_summary$SCIENCE_sd)
)

p3 <- ggplot(stats_long, aes(x = Subject, y = Value, fill = Subject)) +
  geom_col(alpha = 0.8) +
  geom_text(aes(label = round(Value, 1)), vjust = -0.3, 
            fontface = "bold", size = 3.5) +
  facet_wrap(~Metric, scales = "free_y") +
  scale_fill_manual(values = c("MATHS" = "#bfdbfe", 
                                "ENGLISH" = "#bbf7d0", 
                                "SCIENCE" = "#e9d5ff")) +
  labs(title = "Mean & Variability Comparison", y = "Value") +
  theme_minimal() +
  theme(legend.position = "none",
        plot.title = element_text(face = "bold", hjust = 0.5, size = 12))

# Combine
(p1 / (p2 | p3)) +
  plot_annotation(
    title = "Academic Performance: Visual Summary",
    theme = theme(plot.title = element_text(size = 15, face = "bold", hjust = 0.5))
  )
```

```{r}
#| echo: false
# Calculate key stats
maths_mean <- mean(exam_data$MATHS)
english_mean <- mean(exam_data$ENGLISH)
science_mean <- mean(exam_data$SCIENCE)
best_subject <- c("MATHS", "ENGLISH", "SCIENCE")[which.max(c(maths_mean, english_mean, science_mean))]
```

::: {.callout-tip}
## What I Learned

**Distribution shapes**: All three subjects show roughly normal distributions with slight variations. The violin plots reveal the full shape that box plots alone would hide.

**Best performing subject**: `r best_subject` has the highest mean (`r round(max(c(maths_mean, english_mean, science_mean)), 1)` points).

**Variability**: Standard deviations are similar across subjects (~`r round(mean(c(stats_summary$MATHS_sd, stats_summary$ENGLISH_sd, stats_summary$SCIENCE_sd)), 1)` points), indicating consistent spread.

**Diamond markers** = means, **Box** = quartiles, **Dots** = outliers

**My Reflection**: The overlapping density curves suggest students who excel in one subject tend to excel in others - hinting at general academic ability rather than subject-specific talent.
:::

# Grammar of Graphics: Building Blocks

## The Seven Layers

Grammar of Graphics defines **seven grammatical elements**:

1. **Data**: The dataset
2. **Aesthetics**: Visual mappings (x, y, color, size)
3. **Geometrics**: Visual marks (points, bars, lines)
4. **Facets**: Small multiples
5. **Statistics**: Transformations (smoothing, binning)
6. **Coordinates**: The plotting space
7. **Themes**: Non-data styling

Let's explore each layer systematically.

# Layer 1: Data Foundation

## Starting with the Canvas

```{r}
#| code-fold: true
#| code-summary: "Show code"
ggplot(data = exam_data)
```

::: {.callout-note}
## Understanding the Foundation

**What happened**: Created a blank ggplot object with `r n_students` observations loaded.

**What's missing**: No aesthetic mappings yet - ggplot doesn't know what to display.

**My Learning**: This is like preparing a blank canvas before painting. The data is loaded and ready, but we haven't told ggplot *what* to draw or *where* to draw it.
:::

# Layer 2: Aesthetic Mappings

## Mapping Data to Visual Properties

```{r}
#| code-fold: true
#| code-summary: "Show code"
ggplot(data = exam_data, aes(x = MATHS))
```

::: {.callout-note}
## Adding the X-Axis

**What changed**: Mapped MATHS scores to x-axis position.

**Range**: `r min(exam_data$MATHS)` to `r max(exam_data$MATHS)` points (span: `r max(exam_data$MATHS) - min(exam_data$MATHS)` points).

**Still missing**: Geometric objects to visualize the data.

**My Insight**: Aesthetics define *where* data goes, but not *how* it looks. We need geometries next.
:::

# Layer 3: Geometric Objects

## Histogram: Distribution Analysis

### Comparing Bin Choices

::: {.panel-tabset}

## Default (30 bins)

```{r}
#| code-fold: true
#| code-summary: "Show code"
ggplot(exam_data, aes(x = MATHS)) +
  geom_histogram(fill = "#bfdbfe", color = "#1e40af", alpha = 0.8) +
  labs(title = "Default: 30 bins",
       subtitle = paste0("Bin width ≈ ", round((max(exam_data$MATHS) - min(exam_data$MATHS))/30, 1), " points"),
       x = "Mathematics Score", y = "Frequency") +
  theme_minimal()
```

## Optimized (20 bins)

```{r}
#| code-fold: true
#| code-summary: "Show code"
ggplot(exam_data, aes(x = MATHS)) +
  geom_histogram(bins = 20, fill = "#bbf7d0", color = "#065f46", alpha = 0.9) +
  labs(title = "Optimized: 20 bins",
       subtitle = paste0("Bin width ≈ ", round((max(exam_data$MATHS) - min(exam_data$MATHS))/20, 1), " points"),
       x = "Mathematics Score", y = "Frequency") +
  theme_minimal()
```

## Too Few (5 bins)

```{r}
#| code-fold: true
#| code-summary: "Show code"
ggplot(exam_data, aes(x = MATHS)) +
  geom_histogram(bins = 5, fill = "#e9d5ff", color = "#6b21a8", alpha = 0.7) +
  labs(title = "Under-binned: 5 bins",
       subtitle = "Loses important detail",
       x = "Mathematics Score", y = "Frequency") +
  theme_minimal()
```

:::

```{r}
#| echo: false
# Calculate optimal bins
optimal_bins <- ceiling(log2(n_students) + 1)  # Sturges' rule
```

::: {.callout-important}
## Bin Selection Matters

**Statistical recommendation (Sturges' rule)**: `r optimal_bins` bins for n = `r n_students`

**My choice**: 20 bins - balances detail with clarity

**What I discovered**:
- **Too few bins (5)**: Lose distribution shape - looks artificially uniform
- **Too many bins (30)**: Noisy, hard to see overall pattern  
- **Just right (20)**: Clear shape without excessive noise

**My Reflection**: Bin choice is like choosing photo resolution - too low loses detail, too high shows noise. There's a sweet spot that reveals the true pattern.
:::

## Dot Plot: Individual Observations

```{r}
#| code-fold: true
#| code-summary: "Show code"
#| fig-width: 10
#| fig-height: 5
ggplot(exam_data, aes(x = MATHS)) +
  geom_dotplot(binwidth = 2.5, fill = "#f472b6", color = "#831843", 
               dotsize = 0.6) +
  scale_y_continuous(NULL, breaks = NULL) +
  labs(title = "Every Student Counts: Dot Plot of Mathematics Scores",
       subtitle = paste0("Each dot = 1 student | Total: ", n_students, " students"),
       x = "Mathematics Score") +
  theme_minimal() +
  theme(plot.title = element_text(face = "bold", size = 13))
```

::: {.callout-tip}
## Why Dot Plots Are Powerful

**Each dot = one student's story**

**What I see**: 
- Clustering around 60-70 range
- Few students below 40 (need support)
- Some high achievers above 90
- Gaps reveal score ranges with no students

**vs Histogram**: Histogram bins aggregate; dot plot preserves every individual observation

**My Learning**: This visualization makes data **human** - I can see that each dot represents a real student. It's not just statistics, it's `r n_students` individuals with different abilities.
:::

## Bar Chart: Categorical Distribution

```{r}
#| code-fold: true  
#| code-summary: "Show code"
#| fig-width: 10
#| fig-height: 6
exam_data %>%
  count(RACE) %>%
  mutate(pct = n/sum(n) * 100,
         RACE = fct_reorder(RACE, n)) %>%
  ggplot(aes(x = RACE, y = n, fill = n)) +
  geom_col(alpha = 0.85) +
  geom_text(aes(label = paste0(n, " students\n(", round(pct, 1), "%)")),
            hjust = -0.1, size = 3.5, fontface = "bold") +
  scale_fill_gradient(low = "#fce7f3", high = "#ec4899") +
  coord_flip() +
  labs(title = "Student Distribution by Race",
       subtitle = "Horizontal bars make labels readable",
       x = "", y = "Number of Students") +
  theme_minimal() +
  theme(legend.position = "none",
        plot.title = element_text(face = "bold", size = 13))
```

::: {.callout-tip}
## Design Choice: Horizontal Bars

**Why flip coordinates?**

- **Readability**: Long category names don't need rotation  
- **Comparison**: Easier to compare lengths left-to-right  
- **Accessibility**: Natural reading direction

**My Preference**: Always use horizontal bars for categorical data - it respects how humans naturally read.
:::

## Density Curves: Smooth Distributions

### Gender Comparison

```{r}
#| code-fold: true
#| code-summary: "Show code"
#| fig-width: 10
#| fig-height: 6
ggplot(exam_data, aes(x = MATHS, fill = GENDER, color = GENDER)) +
  geom_density(alpha = 0.5, size = 1.3) +
  scale_fill_manual(values = c("Female" = "#fbcfe8", "Male" = "#d8b4fe")) +
  scale_color_manual(values = c("Female" = "#db2777", "Male" = "#a855f7")) +
  labs(title = "Mathematics Performance by Gender",
       subtitle = "Density curves reveal distribution overlap",
       x = "Mathematics Score", y = "Density") +
  theme_minimal() +
  theme(legend.position = "top",
        plot.title = element_text(face = "bold", size = 13))
```

```{r}
#| code-fold: true
#| code-summary: "Show code"
# Statistical test
male_scores <- exam_data$MATHS[exam_data$GENDER == "Male"]
female_scores <- exam_data$MATHS[exam_data$GENDER == "Female"]
t_test <- t.test(male_scores, female_scores)
gender_diff <- abs(mean(male_scores) - mean(female_scores))
```

::: {.callout-important}
## Statistical Analysis

**Visual observation**: Massive overlap between male and female distributions.

**Statistical test**:
- Mean difference: `r round(gender_diff, 2)` points  
- p-value: `r round(t_test$p.value, 4)`  
- Conclusion: `r ifelse(t_test$p.value < 0.05, "Statistically significant", "NOT statistically significant")`

**What this means**:
Even if statistically different, the `r round(gender_diff, 1)`-point gap is only `r round(gender_diff/mean(exam_data$MATHS)*100, 1)`% of the mean - **practically negligible**.

**My Key Learning**: The extensive overlap shows that **individual variation >> gender-based variation**. Gender is NOT a strong predictor of math performance. This challenges common stereotypes.
:::

## Box Plots: Five-Number Summary

### With Statistical Layers

```{r}
#| code-fold: true
#| code-summary: "Show code"
#| fig-width: 10
#| fig-height: 6
ggplot(exam_data, aes(y = MATHS, x = GENDER, fill = GENDER)) +
  geom_violin(alpha = 0.3, trim = FALSE) +
  geom_boxplot(width = 0.3, alpha = 0.7, outlier.shape = 21,
               outlier.fill = "#831843", outlier.size = 3) +
  geom_jitter(width = 0.1, alpha = 0.2, size = 1.5, color = "#831843") +
  stat_summary(fun = mean, geom = "point", shape = 23, 
               size = 5, fill = "red", color = "darkred") +
  scale_fill_manual(values = c("Female" = "#fbcfe8", "Male" = "#d8b4fe")) +
  labs(title = "Mathematics Scores: Layered Visualization",
       subtitle = "Violin + Box + Points + Mean (red diamond)",
       y = "Mathematics Score", x = "") +
  theme_minimal() +
  theme(legend.position = "none",
        plot.title = element_text(face = "bold", size = 13))
```

```{r}
#| echo: false
# Calculate quartiles
quartiles_male <- quantile(male_scores)
quartiles_female <- quantile(female_scores)
```

::: {.callout-tip}
## Layered Visualization Power

**Four layers combined**:

1. **Violin** (background): Full distribution shape
2. **Box plot**: Median, Q1, Q3, whiskers
3. **Jittered points**: Every individual observation  
4. **Red diamond**: Mean value

**Five-number summary**:

| Gender | Min | Q1 | Median | Q3 | Max |
|--------|-----|-----|--------|-----|-----|
| Male | `r round(quartiles_male[1], 1)` | `r round(quartiles_male[2], 1)` | `r round(quartiles_male[3], 1)` | `r round(quartiles_male[4], 1)` | `r round(quartiles_male[5], 1)` |
| Female | `r round(quartiles_female[1], 1)` | `r round(quartiles_female[2], 1)` | `r round(quartiles_female[3], 1)` | `r round(quartiles_female[4], 1)` | `r round(quartiles_female[5], 1)` |

**My Reflection**: Combining multiple geometries gives readers options - they can see summary statistics OR individual points OR distribution shape. Maximum information density.
:::

## Scatter Plots: Relationships

### Mathematics vs English Performance

```{r}
#| code-fold: true
#| code-summary: "Show code"
#| fig-width: 10
#| fig-height: 8
ggplot(exam_data, aes(x = MATHS, y = ENGLISH)) +
  geom_point(alpha = 0.5, size = 3, color = "#db2777") +
  geom_smooth(method = "lm", se = TRUE, color = "#831843", 
              fill = "#fbcfe8", size = 1.2) +
  geom_hline(yintercept = mean(exam_data$ENGLISH), 
             linetype = "dashed", color = "#ec4899", alpha = 0.6) +
  geom_vline(xintercept = mean(exam_data$MATHS), 
             linetype = "dashed", color = "#ec4899", alpha = 0.6) +
  annotate("text", x = mean(exam_data$MATHS), y = 5, 
           label = paste0("Mean MATHS: ", round(mean(exam_data$MATHS), 1)),
           color = "#831843", fontface = "bold", size = 3.5) +
  annotate("text", x = 20, y = mean(exam_data$ENGLISH), 
           label = paste0("Mean ENGLISH: ", round(mean(exam_data$ENGLISH), 1)),
           color = "#831843", fontface = "bold", angle = 90, size = 3.5) +
  labs(title = "Subject Performance Correlation",
       subtitle = "Strong positive relationship suggests interconnected academic skills",
       x = "Mathematics Score", y = "English Score") +
  theme_minimal() +
  theme(plot.title = element_text(face = "bold", size = 13))
```

```{r}
#| code-fold: true
#| code-summary: "Show code"
# Correlation analysis
cor_test <- cor.test(exam_data$MATHS, exam_data$ENGLISH)
lm_model <- lm(ENGLISH ~ MATHS, data = exam_data)
```

::: {.callout-important}
## Correlation Insights

**Pearson correlation**: r = `r round(cor_test$estimate, 3)`  
**R-squared**: `r round(summary(lm_model)$r.squared, 3)` → **`r round(summary(lm_model)$r.squared*100, 1)`% shared variance**

**Regression equation**:  
English = `r round(coef(lm_model)[1], 2)` + `r round(coef(lm_model)[2], 3)` × Maths

**What this means**:
- Every 10-point increase in Maths → `r round(coef(lm_model)[2] * 10, 1)`-point increase in English
- `r round(summary(lm_model)$r.squared*100, 1)`% of English variability explained by Maths performance

**My Discovery**: This strong correlation (r = `r round(cor_test$estimate, 3)`) reveals that academic success is **interconnected**, not subject-specific. Students with strong analytical skills (math) tend to also excel in language skills (English).

**Educational implication**: Focus on general cognitive abilities and study habits rather than subject-specific drilling.
:::

# Layer 4: Facets (Small Multiples)

## Class-Level Performance Breakdown

```{r}
#| code-fold: true
#| code-summary: "Show code"
#| fig-width: 12
#| fig-height: 8
ggplot(exam_data, aes(x = MATHS)) +
  geom_histogram(bins = 15, fill = "#bbf7d0", color = "#065f46", alpha = 0.8) +
  geom_vline(aes(xintercept = mean(MATHS)), 
             color = "red", linetype = "dashed", size = 1) +
  facet_wrap(~ CLASS, ncol = 3) +
  labs(title = "Mathematics Distribution by Class",
       subtitle = "Red line = class mean | Compare shapes and central tendencies",
       x = "Mathematics Score", y = "Frequency") +
  theme_minimal() +
  theme(plot.title = element_text(face = "bold", size = 13),
        strip.text = element_text(face = "bold", size = 10))
```

```{r}
#| code-fold: true
#| code-summary: "Show code"
# Class statistics
class_stats <- exam_data %>%
  group_by(CLASS) %>%
  summarise(
    n = n(),
    mean_maths = mean(MATHS),
    sd_maths = sd(MATHS)
  ) %>%
  arrange(desc(mean_maths))
```

::: {.callout-tip}
## Inter-Class Variation Analysis

**Performance ranking**:

```{r}
#| echo: false
kable(class_stats, digits = 1, 
      col.names = c("Class", "Students (n)", "Mean", "SD"),
      caption = "Class Performance Summary")
```

**Key observations**:

1. **Best class**: `r class_stats$CLASS[1]` (μ = `r round(class_stats$mean_maths[1], 1)`)
2. **Weakest class**: `r class_stats$CLASS[nrow(class_stats)]` (μ = `r round(class_stats$mean_maths[nrow(class_stats)], 1)`)
3. **Gap**: `r round(class_stats$mean_maths[1] - class_stats$mean_maths[nrow(class_stats)], 1)` points

**My Learning**: Facets allow simultaneous comparison of multiple groups. The variation suggests teaching quality or class composition matters. This warrants investigation - are high-performing students clustered in certain classes?
:::

# Layer 5: Statistical Transformations

## Adding Regression Lines

::: {.panel-tabset}

## Linear Model

```{r}
#| code-fold: true
#| code-summary: "Show code"
ggplot(exam_data, aes(x = MATHS, y = ENGLISH)) +
  geom_point(alpha = 0.4, color = "#db2777", size = 2) +
  geom_smooth(method = "lm", se = TRUE, color = "#831843", 
              fill = "#fbcfe8", size = 1.2) +
  labs(title = "Linear Regression: Parametric Assumption",
       subtitle = paste0("R² = ", round(summary(lm_model)$r.squared, 3)),
       x = "Mathematics", y = "English") +
  theme_minimal() +
  theme(plot.title = element_text(face = "bold", size = 12))
```

## LOESS Smoothing

```{r}
#| code-fold: true
#| code-summary: "Show code"
ggplot(exam_data, aes(x = MATHS, y = ENGLISH)) +
  geom_point(alpha = 0.4, color = "#db2777", size = 2) +
  geom_smooth(method = "loess", se = TRUE, color = "#831843", 
              fill = "#fbcfe8", size = 1.2) +
  labs(title = "LOESS Smoothing: Non-Parametric Flexibility",
       subtitle = "Captures local patterns without assuming linearity",
       x = "Mathematics", y = "English") +
  theme_minimal() +
  theme(plot.title = element_text(face = "bold", size = 12))
```

:::

::: {.callout-note}
## Linear vs LOESS

**Linear (lm)**:
- Assumes straight-line relationship
- Simple, interpretable  
- Good when relationship is truly linear

**LOESS (locally weighted)**:
- No parametric assumptions
- Flexible, captures curves
- Better for exploring non-linear patterns

**For our data**: Both methods produce similar fits → relationship is genuinely linear. This validates using simple regression.

**My Insight**: When linear and LOESS agree, you can trust the simpler model. When they diverge, investigate why.
:::

# Layer 6: Coordinate Systems

## Equal Scales for Honest Comparison

::: {.panel-tabset}

## Misleading: Default Scales

```{r}
#| code-fold: true
#| code-summary: "Show code"
ggplot(exam_data, aes(x = MATHS, y = ENGLISH)) +
  geom_point(alpha = 0.4, color = "#db2777") +
  geom_smooth(method = "lm", color = "#831843", fill = "#fbcfe8") +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", 
              color = "red", alpha = 0.5) +
  labs(title = "MISLEADING: Unequal Scales",
       subtitle = "Regression line angle is distorted",
       x = "Mathematics Score", y = "English Score") +
  theme_minimal() +
  theme(plot.title = element_text(face = "bold", size = 12))
```

## Honest: Equal Scales

```{r}
#| code-fold: true
#| code-summary: "Show code"
ggplot(exam_data, aes(x = MATHS, y = ENGLISH)) +
  geom_point(alpha = 0.4, color = "#db2777") +
  geom_smooth(method = "lm", color = "#831843", fill = "#fbcfe8") +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", 
              color = "red", alpha = 0.5) +
  coord_equal(xlim = c(0, 100), ylim = c(0, 100)) +
  labs(title = "HONEST: Equal Scales (0-100)",
       subtitle = "True representation of relationship strength",
       x = "Mathematics Score", y = "English Score") +
  theme_minimal() +
  theme(plot.title = element_text(face = "bold", size = 12))
```

:::

::: {.callout-warning}
## Visual Deception Alert

**Red dashed line** = perfect 1:1 relationship (slope = 1)

**With equal scales**: You can see the regression line's true angle relative to 1:1

**Why this matters**: 
- Unequal scales can make weak correlations look strong (or vice versa)
- Equal scales provide honest visual representation
- Critical for comparing similar measurements

**My Learning**: This is how visualizations can mislead. Always question axis scales when evaluating relationships. Insist on equal scales when comparing like measurements.
:::

# Layer 7: Themes

## Professional Styling

::: {.panel-tabset}

## theme_minimal() 

```{r}
#| code-fold: true
#| code-summary: "Show code"
ggplot(exam_data, aes(x = RACE, fill = RACE)) +
  geom_bar(alpha = 0.8) +
  scale_fill_manual(values = c("#fce7f3", "#fbcfe8", "#f9a8d4", "#f472b6")) +
  coord_flip() +
  theme_minimal() +
  labs(title = "theme_minimal() - Modern & Clean",
       x = "", y = "Count") +
  theme(legend.position = "none",
        plot.title = element_text(face = "bold", size = 12))
```

## theme_classic()

```{r}
#| code-fold: true
#| code-summary: "Show code"
ggplot(exam_data, aes(x = RACE, fill = RACE)) +
  geom_bar(alpha = 0.8) +
  scale_fill_manual(values = c("#fce7f3", "#fbcfe8", "#f9a8d4", "#f472b6")) +
  coord_flip() +
  theme_classic() +
  labs(title = "theme_classic() - Publication Ready",
       x = "", y = "Count") +
  theme(legend.position = "none",
        plot.title = element_text(face = "bold", size = 12))
```

## theme_bw()

```{r}
#| code-fold: true
#| code-summary: "Show code"
ggplot(exam_data, aes(x = RACE, fill = RACE)) +
  geom_bar(alpha = 0.8) +
  scale_fill_manual(values = c("#fce7f3", "#fbcfe8", "#f9a8d4", "#f472b6")) +
  coord_flip() +
  theme_bw() +
  labs(title = "theme_bw() - Black & White Professional",
       x = "", y = "Count") +
  theme(legend.position = "none",
        plot.title = element_text(face = "bold", size = 12))
```

:::

::: {.callout-tip}
## Theme Selection Guide

| Theme | Use Case | Rating |
|-------|----------|--------|
| **minimal** | Web, dashboards, modern reports | Excellent |
| **classic** | Academic papers, journals | Very Good |
| **bw** | Print publications, formal docs | Very Good |

**My preference**: `theme_minimal()` - clean, doesn't distract from data, works everywhere.

**Professional tip**: Consistency matters more than which theme you choose. Pick one and use it throughout your document.
:::

# Comprehensive Subject Analysis

## Inter-Subject Correlations

```{r}
#| code-fold: true
#| code-summary: "Show code"
#| fig-width: 12
#| fig-height: 4

# Calculate correlations
cor_matrix <- cor(exam_data[, c("MATHS", "ENGLISH", "SCIENCE")])

# Create correlation plot data
cor_data <- expand.grid(
  Var1 = c("MATHS", "ENGLISH", "SCIENCE"),
  Var2 = c("MATHS", "ENGLISH", "SCIENCE")
) %>%
  mutate(Correlation = c(cor_matrix))

# Correlation heatmap
p1 <- ggplot(cor_data, aes(x = Var1, y = Var2, fill = Correlation)) +
  geom_tile(color = "white", size = 1.5) +
  geom_text(aes(label = round(Correlation, 2)), size = 5, fontface = "bold") +
  scale_fill_gradient2(low = "#fce7f3", mid = "#f9a8d4", high = "#ec4899",
                       midpoint = 0.5, limits = c(0, 1)) +
  labs(title = "Correlation Matrix", x = "", y = "") +
  theme_minimal() +
  theme(plot.title = element_text(face = "bold", hjust = 0.5, size = 12))

# Scatter plot matrix sample
p2 <- ggplot(exam_data, aes(x = MATHS, y = SCIENCE)) +
  geom_point(alpha = 0.4, color = "#db2777", size = 2) +
  geom_smooth(method = "lm", color = "#831843", fill = "#fbcfe8") +
  labs(title = "MATHS vs SCIENCE") +
  theme_minimal() +
  theme(plot.title = element_text(face = "bold", size = 11))

p3 <- ggplot(exam_data, aes(x = ENGLISH, y = SCIENCE)) +
  geom_point(alpha = 0.4, color = "#db2777", size = 2) +
  geom_smooth(method = "lm", color = "#831843", fill = "#fbcfe8") +
  labs(title = "ENGLISH vs SCIENCE") +
  theme_minimal() +
  theme(plot.title = element_text(face = "bold", size = 11))

(p1 | (p2 / p3))
```

```{r}
#| echo: false
cor_me <- cor_matrix["MATHS", "ENGLISH"]
cor_ms <- cor_matrix["MATHS", "SCIENCE"]
cor_es <- cor_matrix["ENGLISH", "SCIENCE"]
avg_cor <- mean(c(cor_me, cor_ms, cor_es))
```

::: {.callout-important}
## Key Discovery: Interconnected Learning

**All correlations are strong positive**:

- MATHS ↔ ENGLISH: r = `r round(cor_me, 3)`
- MATHS ↔ SCIENCE: r = `r round(cor_ms, 3)`  
- ENGLISH ↔ SCIENCE: r = `r round(cor_es, 3)`

**Average correlation**: `r round(avg_cor, 3)` → **Strong interconnection**

**What this reveals**:
Academic abilities are NOT isolated silos. Students who excel in one subject tend to excel in all subjects. This suggests:

1. **General cognitive ability** > subject-specific talent
2. **Study habits & motivation** transfer across subjects
3. **Intervention in one subject** may boost overall performance

**My Educational Insight**: Rather than subject-specific remediation, focus on building general learning skills, critical thinking, and study strategies that benefit all subjects.
:::

# Final Reflections

## My Learning Journey

::: welcome-section
## What I Learned After This Journey

**Technical Skills**:
- Seven layers of Grammar of Graphics  
- Choosing appropriate geometric objects  
- Visual vs statistical significance  
- Honest visualization practices

**Analytical Insights**:
- Gender differences are statistically tiny and practically irrelevant  
- Subject performances are strongly interconnected (avg r = `r round(avg_cor, 2)`)  
- Individual variation >> group differences  
- Bin selection and scale choices dramatically affect interpretation

**Most Important Lesson**:
Visualization is not just about making pretty pictures - it's about **revealing truth** and **challenging assumptions**. The gender analysis taught me that what we *expect* to find (gender differences) may not be what the data *actually shows* (minimal differences, huge overlap).

**Next Steps**:
Apply these principles to more complex datasets and tackle multivariate visualizations with confidence.
:::

# References

- Hadley Wickham (2023) [**ggplot2: Elegant Graphics for Data Analysis**](https://ggplot2-book.org/)
- Winston Chang (2013) [**R Graphics Cookbook**](https://r-graphics.org/)
- Healy, Kieran (2019) [**Data Visualization: A practical introduction**](https://socviz.co/)
