---
title: "Hands-on Exercise 3a"
subtitle: "Interactive Data Visualisation — ggiraph, plotly & crosstalk"
author: "Thi Thanh Van Tran"
date: "April 22, 2026"
date-modified: "April 22, 2026"
format:
  html:
    toc: true
    number-sections: true
    number-depth: 4
    code-fold: true
    code-summary: "Show code"
    code-tools: false
    code-copy: true
    css: styles.css
    include-after-body: toc-tracker.html
    theme: cosmo
    page-layout: article
execute:
  eval: true
  echo: true
  warning: false
  freeze: false
knitr:
  opts_chunk:
    fig.width: 10
    fig.height: 6
    dpi: 120
    fig.align: "center"
    out.width: "100%"
---

::: welcome-section
# Interactive Data Visualisation: Mastering ggiraph, plotly & crosstalk

Welcome! This hands-on exercise covers **three complementary tools** for interactive visualisation in R:

1. **ggiraph** — extends ggplot2 with hover tooltips, `data_id` linking, and `onclick` actions
2. **plotly** — full-featured interactive charts via `plot_ly()` and `ggplotly()`
3. **crosstalk** — cross-widget interactions linking plots to interactive tables
:::

---

## Learning Outcome {#learning-outcome}

In this hands-on exercise, you will learn how to create interactive data visualisations by using functions provided by the **ggiraph** and **plotly** packages. Specifically, we will explore:

- How to add hover tooltips to ggplot2 charts using `geom_*_interactive()`
- How to link multiple charts with shared `data_id` keys for coordinated highlighting
- How to deploy `onclick` actions that open external URLs from chart elements
- How to convert ggplot2 objects to interactive plotly widgets with `ggplotly()`
- How to build linked brushing dashboards using the crosstalk SharedData pattern

---

## Getting Started {#getting-started}

First, check, install and launch the following R packages:

| Package | Purpose |
|---------|---------|
| **ggiraph** | Makes ggplot2 graphics interactive via tooltip, data\_id and onclick |
| **plotly** | Interactive web graphics via `plot_ly()` and `ggplotly()` |
| **DT** | R interface to JavaScript DataTables for interactive HTML tables |
| **tidyverse** | Modern R packages for data science (ggplot2, dplyr, readr, etc.) |
| **patchwork** | Combines multiple ggplot2 graphs; enables coordinated ggiraph views |
| **readxl** | Imports Excel workbooks |
| **viridis** | Perceptually uniform colour palettes |
| **scales** | Formatting helpers for axis labels and tooltips |
| **crosstalk** | Cross-widget linked brushing and filtering between htmlwidgets |

```{r}
#| label: load-packages
pacman::p_load(ggiraph, plotly, DT, tidyverse, readxl,
               patchwork, viridis, scales, crosstalk)
```

::: {.callout-note}
## What I Learned
`pacman::p_load()` installs any missing package before loading — one line replaces multiple `install.packages()` + `library()` calls. This keeps setup code portable across machines without manual dependency management.
:::

---

## Importing Data {#importing-data}

The *Data* worksheet from the *GlobalPopulation* Excel workbook is used throughout this exercise. It contains country-level demographic records including: `Country`, `Continent`, `Year`, `Population`, `Young` (% young-age population) and `Old` (% elderly population).

```{r}
#| label: import-data
pop_data <- read_xls("data/GlobalPopulation.xls", sheet = "Data") %>%
  mutate(across(where(is.character), as.factor))
```

::: {.callout-note}
## What I Learned
`read_xls()` from **readxl** handles Excel files natively — no CSV conversion needed. `mutate(across(where(is.character), as.factor))` is the modern dplyr idiom for bulk type conversion: `where(is.character)` acts as a column selector predicate, so any new character column added later is also automatically converted. This is more robust than listing column names manually.
:::

---

## Visual Data Exploration {#data-exploration}

Before building interactive charts, it is essential to understand the structure and distribution of the data. Visual exploration helps identify patterns, skewness, and grouping variables that will guide interactive design choices.

::: {.panel-tabset}

### Dataset Overview

```{r}
#| label: data-overview
#| fig.height: 9

n_records  <- nrow(pop_data)
n_complete <- sum(complete.cases(pop_data))

var_types <- data.frame(Type = sapply(pop_data, function(x) class(x)[1])) %>%
  count(Type)

p_sample <- ggplot(data.frame(x = 1, y = 1), aes(x, y)) +
  geom_tile(fill = "#bfdbfe", alpha = 0.6, width = 0.9, height = 0.9) +
  geom_text(label = paste0("n = ", format(n_records, big.mark = ","), "\nrecords"),
            size = 9, fontface = "bold", colour = "#1e3a5f") +
  labs(title = "Sample Size") +
  theme_void() +
  theme(plot.title = element_text(hjust = 0.5, face = "bold", size = 14))

p_types <- ggplot(var_types, aes(x = reorder(Type, -n), y = n, fill = Type)) +
  geom_col(alpha = 0.85, width = 0.6) +
  geom_text(aes(label = n), vjust = -0.5, fontface = "bold", size = 5) +
  scale_fill_manual(values = c("factor"  = "#a5b4fc",
                               "integer" = "#bfdbfe",
                               "numeric" = "#93c5fd")) +
  labs(title = "Variable Types", x = "Type", y = "Count") +
  ylim(0, max(var_types$n) * 1.25) +
  theme_minimal() +
  theme(plot.title = element_text(hjust = 0.5, face = "bold", size = 14),
        legend.position = "none")

missing_df <- reshape2::melt(is.na(pop_data))

p_missing <- ggplot(missing_df, aes(x = Var2, y = Var1, fill = value)) +
  geom_tile(colour = "white", linewidth = 0.05) +
  scale_fill_manual(values = c("FALSE" = "#10b981", "TRUE" = "#ef4444"),
                    labels = c("Complete", "Missing"), name = "Status") +
  labs(title = "Missing Value Heatmap", x = "Variables", y = "Observations") +
  theme_minimal() +
  theme(plot.title    = element_text(hjust = 0.5, face = "bold", size = 14),
        axis.text.y   = element_blank(),
        axis.ticks.y  = element_blank(),
        axis.text.x   = element_text(angle = 45, hjust = 1),
        panel.grid    = element_blank())

(p_sample | p_types) / p_missing +
  plot_annotation(
    title = "GlobalPopulation Dataset Overview",
    theme = theme(plot.title = element_text(size = 16, face = "bold", hjust = 0.5))
  ) +
  plot_layout(heights = c(1, 1.4))
```

::: {.callout-note}
## What I Learned
The overview reveals that the dataset is 100% complete — an important confirmation before building interactive charts, since missing values can cause silent tooltip failures or blank hover panels. The variable type breakdown confirms that `Country` and `Continent` are factors (suitable for `data_id` grouping) while `Young`, `Old`, and `Population` are numeric (suitable for axis mappings and tooltip statistics).
:::

### Categorical Distributions

```{r}
#| label: cat-distributions
#| fig.height: 9

cat_vars      <- names(pop_data)[sapply(pop_data, is.factor)]
pastel_colors <- c("#bfdbfe","#a5b4fc","#bbf7d0","#fde68a",
                   "#fecaca","#fbcfe8","#a7f3d0","#ddd6fe")

plot_list <- lapply(cat_vars, function(var) {
  plot_data <- pop_data %>%
    group_by(.data[[var]]) %>%
    summarise(n = n(), .groups = "drop") %>%
    mutate(pct = n / sum(n) * 100) %>%
    arrange(desc(n))

  n_cats    <- nrow(plot_data)
  fills     <- rep_len(pastel_colors, n_cats)
  title_sfx <- ""
  if (n_cats > 20) { plot_data <- slice_head(plot_data, n = 20); title_sfx <- " (Top 20)" }

  ggplot(plot_data, aes(x = reorder(.data[[var]], -n), y = n, fill = .data[[var]])) +
    geom_col(alpha = 0.9, width = 0.7, colour = "white") +
    geom_text(aes(label = paste0(n, "\n(", round(pct, 1), "%)")),
              vjust = -0.2, size = 2.5, fontface = "bold", colour = "#1e3a5f") +
    scale_fill_manual(values = fills) +
    labs(title = paste(var, "Distribution", title_sfx), y = "Count", x = "") +
    ylim(0, max(plot_data$n) * 1.18) +
    theme_minimal(base_size = 10) +
    theme(plot.title      = element_text(hjust = 0.5, face = "bold", size = 11),
          legend.position = "none",
          axis.text.x     = element_text(angle = 45, hjust = 1, size = 7))
})

wrap_plots(plot_list, ncol = 2) +
  plot_annotation(
    title = "Categorical Variable Distributions",
    theme = theme(plot.title = element_text(size = 13, face = "bold", hjust = 0.5))
  )
```

::: {.callout-note}
## What I Learned
Africa has the most country-records in the dataset while Oceania has the fewest. This imbalance matters for interactive design: continent-level `data_id` grouping will produce dense hover clusters for Africa and sparse ones for Oceania. Understanding this distribution before coding helps anticipate which interactions will be visually clear versus confusing.
:::

### Numeric Distributions

```{r}
#| label: numeric-distributions
#| fig.height: 8

latest_year <- max(pop_data$Year, na.rm = TRUE)
latest_snap <- pop_data %>% filter(Year == latest_year)

ph1 <- ggplot(latest_snap, aes(x = Old)) +
  geom_histogram(bins = 30, fill = "#3b82f6", colour = "white", alpha = 0.85) +
  labs(title = paste("% Aged —", latest_year), x = "% Aged", y = "Count") +
  theme_minimal()

ph2 <- ggplot(latest_snap, aes(x = Young)) +
  geom_histogram(bins = 30, fill = "#ec4899", colour = "white", alpha = 0.85) +
  labs(title = paste("% Young —", latest_year), x = "% Young", y = "Count") +
  theme_minimal()

ph3 <- ggplot(latest_snap, aes(x = Population / 1e6)) +
  geom_histogram(bins = 30, fill = "#10b981", colour = "white", alpha = 0.85) +
  labs(title = paste("Population (M) —", latest_year),
       x = "Population (Millions)", y = "Count") +
  theme_minimal()

pb1 <- ggplot(latest_snap, aes(x = Continent, y = Old, fill = Continent)) +
  geom_boxplot(alpha = 0.7, outlier.size = 1.5) +
  scale_fill_viridis_d(option = "plasma") +
  labs(title = "% Aged by Continent", x = "", y = "% Aged") +
  theme_minimal() +
  theme(legend.position = "none",
        axis.text.x = element_text(angle = 30, hjust = 1))

(ph1 | ph2 | ph3) / pb1 +
  plot_annotation(
    title = "Numeric Variable Distributions",
    theme = theme(plot.title = element_text(size = 14, face = "bold", hjust = 0.5))
  ) +
  plot_layout(heights = c(1, 1.3))
```

::: {.callout-note}
## What I Learned
Population is highly right-skewed — a handful of very populous countries dominate the scale. This explains why interactive charts use `sizes = c(2, 100)` in plotly or `scale_size(range = c(2, 12))` in ggplot2 to compress the mapping and keep small countries visible. The boxplot confirms that Europe has the highest median ageing percentage — a finding that interactive hover tooltips will let users drill into at the country level.
:::

### Variable Relationships

```{r}
#| label: variable-relationships
#| fig.height: 9

r1 <- ggplot(latest_snap, aes(x = Old, y = Young, colour = Continent)) +
  geom_point(alpha = 0.6, size = 2.5) +
  scale_colour_viridis_d(option = "turbo") +
  labs(title = "% Aged vs % Young (latest year)",
       subtitle = "Strong negative correlation — the demographic trade-off",
       x = "% Aged", y = "% Young") +
  theme_minimal()

r2 <- pop_data %>%
  group_by(Continent, Year) %>%
  summarise(Mean_Old = mean(Old, na.rm = TRUE), .groups = "drop") %>%
  ggplot(aes(x = Year, y = Mean_Old, colour = Continent)) +
  geom_line(linewidth = 0.9) +
  geom_point(size = 1.5) +
  scale_colour_brewer(palette = "Set1") +
  labs(title = "Mean % Aged Over Time by Continent",
       subtitle = "All continents ageing — rate differs",
       x = "Year", y = "Mean % Aged") +
  theme_minimal()

r1 / r2 +
  plot_annotation(
    title = "Key Data Relationships",
    theme = theme(plot.title = element_text(size = 14, face = "bold", hjust = 0.5))
  )
```

::: {.callout-note}
## What I Learned
The strong negative correlation between % Aged and % Young is the central analytical story of this dataset. Understanding it upfront directly motivates the interactive design choices ahead: `data_id = Continent` for group highlighting (to compare continental trajectories), and tooltips showing both Old and Young values simultaneously (to reveal the demographic trade-off at country level in a single hover).
:::

:::

---

## Interactive Data Visualisation — ggiraph Methods {#ggiraph-methods}

::: {.callout-tip}
## Core Concept
**ggiraph** is an htmlwidget and ggplot2 extension. It makes ggplot graphics interactive via three aesthetics added to `geom_*_interactive()` functions:

- **`tooltip`** — text shown when mouse hovers over an element
- **`data_id`** — key that links elements for group highlighting
- **`onclick`** — JavaScript string executed on click

Workflow: replace `geom_*()` → `geom_*_interactive()`, add interactive aesthetics inside `aes()`, then wrap the ggplot object in `girafe()`.
:::

### Tooltip Effect with `tooltip` Aesthetic {#tooltip-basic}

Below is a typical code pattern for an interactive statistical graph using **ggiraph**. Two steps are involved: first build a ggplot object with an interactive geom; then pass it to `girafe()` to generate an SVG widget displayed on the HTML page.

```{r}
#| label: tooltip-basic

p <- pop_data %>%
  filter(Year == max(Year, na.rm = TRUE)) %>%
  ggplot(aes(x = Old, y = Young)) +
  geom_point_interactive(
    aes(tooltip = Country),
    size = 3, alpha = 0.7, colour = "#1e40af"
  ) +
  labs(title  = "% Aged vs % Young — Basic Tooltip (hover for country name)",
       subtitle = paste("Latest year:", max(pop_data$Year, na.rm = TRUE)),
       x = "% Aged", y = "% Young") +
  theme_minimal() +
  theme(plot.title = element_text(face = "bold", size = 13))

girafe(
  ggobj      = p,
  width_svg  = 10,
  height_svg = 10 * 0.618
)
```

By hovering the mouse pointer on any data point, the country name is displayed. The only change from a standard ggplot is replacing `geom_point()` with `geom_point_interactive()` and adding `tooltip = Country` inside `aes()`, then wrapping in `girafe()`.

::: {.callout-note}
## What I Learned
The two-step pattern — interactive geom → `girafe()` — is the mental model for all ggiraph work. The SVG output embeds directly in HTML reports with no extra server required. This is the cleanest entry point: one swapped geom + one wrapper, and you have immediate hover feedback. Critically, the tooltip must reference a column that already exists in the data frame — you cannot compute values inline inside `aes()` the way you can with `colour =` or `size =`.
:::

---

### Displaying Multiple Information on Tooltip {#tooltip-multi}

Tooltip content can be customised by constructing a character column that combines multiple fields:

```{r}
#| label: tooltip-multi

latest_multi <- pop_data %>%
  filter(Year == max(Year, na.rm = TRUE)) %>%
  mutate(tooltip_text = paste0(
    "Country = ",    Country,
    "\nContinent = ", Continent,
    "\n% Aged = ",   round(Old, 1),   "%",
    "\n% Young = ",  round(Young, 1), "%"
  ))

p <- ggplot(data = latest_multi, aes(x = Old, y = Young)) +
  geom_point_interactive(
    aes(tooltip = tooltip_text),
    size = 3, alpha = 0.7, colour = "#059669"
  ) +
  labs(title = "% Aged vs % Young — Multi-field Tooltip",
       x = "% Aged", y = "% Young") +
  theme_minimal() +
  theme(plot.title = element_text(face = "bold", size = 13))

girafe(
  ggobj      = p,
  width_svg  = 10,
  height_svg = 10 * 0.618
)
```

The first `mutate()` call creates a new `tooltip_text` column. At the same time, it populates text from `Country`, `Continent`, `Old`, and `Young` fields into the newly created field, which is then passed to `aes(tooltip = tooltip_text)`.

By hovering on a data point, Country, Continent, % Aged, and % Young are all displayed together.

::: {.callout-note}
## What I Learned
Pre-computing a tooltip column with `mutate()` before the ggplot call keeps `aes()` clean and makes complex tooltips easy to debug — you can inspect the tooltip column directly with `head()`. Using explicit field labels (e.g., `"Country = "`) makes tooltips self-explanatory for end users who may not recognise which axis maps to which variable.
:::

---

## Interactivity — Customisation {#interactivity-customisation}

### Customising Tooltip Style {#tooltip-style}

`opts_tooltip()` inside `girafe()` applies CSS to the tooltip container, allowing full visual control over its appearance:

```{r}
#| label: tooltip-style

tooltip_css <- "background-color:#1e293b; color:#f8fafc; font-weight:bold;
                padding:8px 12px; border-radius:6px; font-size:13px;"

latest_styled <- pop_data %>%
  filter(Year == max(Year, na.rm = TRUE)) %>%
  mutate(tooltip_text = paste0(
    "Country = ",    Country,
    "\nContinent = ", Continent
  ))

p <- ggplot(data = latest_styled, aes(x = Old, y = Young)) +
  geom_point_interactive(
    aes(tooltip = tooltip_text),
    size = 3, alpha = 0.7, colour = "#7c3aed"
  ) +
  labs(title = "% Aged vs % Young — Styled Tooltip (dark background)",
       x = "% Aged", y = "% Young") +
  theme_minimal() +
  theme(plot.title = element_text(face = "bold", size = 13))

girafe(
  ggobj      = p,
  width_svg  = 10,
  height_svg = 10 * 0.618,
  options    = list(
    opts_tooltip(css = tooltip_css)
  )
)
```

The `tooltip_css` string is a standard CSS declaration. `opts_tooltip(css = ...)` passes it to the SVG tooltip element. The same CSS properties available in web development — `background-color`, `color`, `font-weight`, `border-radius` — work here. Refer to [Customizing girafe objects](https://www.ardata.fr/ggiraph-book/customize.html) for a full list of options.

::: {.callout-note}
## What I Learned
Tooltip CSS is applied at the `girafe()` level, not inside `aes()` — this cleanly separates data mapping from visual styling. Dark backgrounds with light text (`#1e293b` / `#f8fafc`) immediately distinguish the tooltip from the chart background, improving readability regardless of the chart's colour scheme. This mirrors best practices from web UI design applied directly to R visualisations.
:::

---

### Displaying Statistics on Tooltip {#tooltip-stats}

A function can compute derived statistics and embed them in the tooltip using `after_stat()`:

```{r}
#| label: tooltip-stats

tooltip_fn <- function(y, ymax, accuracy = 0.01) {
  mean_val <- scales::number(y, accuracy = accuracy)
  sem_val  <- scales::number(ymax - y, accuracy = accuracy)
  paste("Mean % Aged:", mean_val, "+/-", sem_val)
}

latest_stat <- pop_data %>% filter(Year == max(Year, na.rm = TRUE))

gg_bar <- ggplot(data = latest_stat, aes(x = Continent)) +
  stat_summary(
    aes(y       = Old,
        tooltip = after_stat(tooltip_fn(y, ymax))),
    fun.data = "mean_se",
    geom     = GeomInteractiveCol,
    fill     = "#93c5fd"
  ) +
  stat_summary(
    aes(y = Old),
    fun.data = mean_se,
    geom     = "errorbar",
    width    = 0.2, linewidth = 0.4
  ) +
  labs(title    = "Mean % Aged by Continent — Statistics on Tooltip",
       subtitle  = "Hover each bar to see Mean ± SEM",
       x = "Continent", y = "Mean % Aged") +
  theme_minimal() +
  theme(plot.title  = element_text(face = "bold", size = 13),
        axis.text.x = element_text(angle = 30, hjust = 1))

girafe(ggobj = gg_bar, width_svg = 10, height_svg = 10 * 0.618)
```

`after_stat()` defers evaluation of `tooltip_fn(y, ymax)` until `stat_summary` has computed the summary statistics — meaning `y` and `ymax` refer to the *computed* mean and upper bound (mean + SE), not raw data columns. `GeomInteractiveCol` is the interactive version of `GeomCol`, enabling the tooltip on bar charts.

::: {.callout-note}
## What I Learned
`after_stat()` is the bridge between statistical transformations and interactive aesthetics. Without it, computed values like `y` (mean) and `ymax` (mean + SE) would not be accessible inside `aes()`. This pattern is essential whenever you want to show computed statistics — not raw values — on hover. For demographic data, showing Mean ± SEM on hover turns a simple bar chart into a rigorous statistical summary without cluttering the visual with permanent annotations.
:::

---

### Hover Effect with `data_id` Aesthetic {#data-id}

`data_id` is the second interactive feature of ggiraph. Elements sharing the same `data_id` value are highlighted together on hover:

```{r}
#| label: data-id-basic

latest_id <- pop_data %>% filter(Year == max(Year, na.rm = TRUE))

p <- ggplot(data = latest_id, aes(x = Old, y = Young, colour = Continent)) +
  geom_point_interactive(
    aes(data_id = Continent),
    size = 3, alpha = 0.7
  ) +
  scale_colour_viridis_d(option = "plasma") +
  labs(title = "Hover to highlight all countries in the same continent",
       x = "% Aged", y = "% Young") +
  theme_minimal() +
  theme(plot.title = element_text(face = "bold", size = 13))

girafe(
  ggobj      = p,
  width_svg  = 10,
  height_svg = 10 * 0.618
)
```

Elements associated with a `data_id` (i.e. `Continent`) will be highlighted upon mouse over. The default value of the hover CSS is `hover_css = "fill:orange;"`. Note that `data_id` and `tooltip` are independent: you can use either or both.

::: {.callout-note}
## What I Learned
`data_id` is the mechanism that makes ggiraph's linking magic possible. Assigning the same key to multiple elements (all points from Asia, for example) causes them to respond as a group. This transforms a simple hover into a powerful grouping tool — the viewer can immediately see all members of a continent cluster without needing colour legend lookup or filtering. This is the foundation for coordinated multiple views later.
:::

---

### Styling the Hover Effect {#hover-style}

CSS codes inside `opts_hover()` and `opts_hover_inv()` change the highlighting colour and dim non-hovered elements:

```{r}
#| label: hover-style

latest_hover <- pop_data %>% filter(Year == max(Year, na.rm = TRUE))

p <- ggplot(data = latest_hover, aes(x = Old, y = Young, colour = Continent)) +
  geom_point_interactive(
    aes(data_id = Continent),
    size = 3, alpha = 0.7
  ) +
  scale_colour_viridis_d(option = "plasma") +
  labs(title = "Styled hover: hovered continent bold, others dimmed",
       x = "% Aged", y = "% Young") +
  theme_minimal() +
  theme(plot.title = element_text(face = "bold", size = 13))

girafe(
  ggobj      = p,
  width_svg  = 10,
  height_svg = 10 * 0.618,
  options    = list(
    opts_hover(css     = "fill:#202020; stroke:#202020;"),
    opts_hover_inv(css = "opacity:0.15;")
  )
)
```

Different from the previous example, here the CSS customisation requests are encoded directly inside `opts_hover()` and `opts_hover_inv()`. `opts_hover()` defines the CSS for the *hovered* element; `opts_hover_inv()` applies to all *other* elements.

::: {.callout-note}
## What I Learned
The `opts_hover_inv()` pattern — dim everything else — is far more effective than highlighting the active element alone. The contrast between `opacity:0.15` (background countries) and the bold dark fill (hovered continent) creates immediate visual hierarchy with only two CSS lines. Applied to demographic data, this lets the viewer instantly isolate one continent's position in the % Aged vs % Young space without losing context of the full distribution.
:::

---

### Combining Tooltip and Hover Effect {#tooltip-hover}

There are times we want to combine tooltip and hover effect on the interactive statistical graph:

```{r}
#| label: tooltip-hover-combined

latest_combo <- pop_data %>%
  filter(Year == max(Year, na.rm = TRUE)) %>%
  mutate(tooltip_text = paste0(
    "Country: ",     Country,
    "\nContinent: ", Continent,
    "\n% Aged: ",    round(Old, 1),   "%",
    "\n% Young: ",   round(Young, 1), "%",
    "\nPopulation: ", format(Population, big.mark = ",")
  ))

p <- ggplot(data = latest_combo, aes(x = Old, y = Young, colour = Continent)) +
  geom_point_interactive(
    aes(tooltip = tooltip_text,
        data_id = Continent),
    size = 3, alpha = 0.7
  ) +
  scale_colour_viridis_d(option = "turbo") +
  labs(title = "Combined: tooltip shows country detail, hover highlights continent",
       x = "% Aged", y = "% Young") +
  theme_minimal() +
  theme(plot.title = element_text(face = "bold", size = 13))

girafe(
  ggobj      = p,
  width_svg  = 10,
  height_svg = 10 * 0.618,
  options    = list(
    opts_hover(css     = "fill:#202020; stroke:#202020;"),
    opts_hover_inv(css = "opacity:0.15;")
  )
)
```

Elements associated with a `data_id` (i.e. `Continent`) will be highlighted upon mouse over. At the same time, the tooltip will show the individual country's full demographic profile.

::: {.callout-note}
## What I Learned
Combining tooltip and `data_id` delivers two analytical layers at once — **individual context** (which country, what values) via tooltip, and **group context** (which continent, which cluster) via hover highlighting. For hierarchical datasets like country-within-continent, this layered interactivity matches the natural analytical workflow: first notice a cluster, then explore specific countries within it. This is impossible to achieve with static charts.
:::

---

### Click Effect with `onclick` {#onclick}

`onclick` argument of ggiraph provides hotlink interactivity on the web. The code chunk below shows an example:

```{r}
#| label: onclick

latest_click <- pop_data %>%
  filter(Year == max(Year, na.rm = TRUE)) %>%
  mutate(
    onclick_js   = paste0(
      "window.open('https://www.google.com/search?q=",
      URLencode(paste(as.character(Country), "population demographics")),
      "')"
    ),
    tooltip_text = paste0(as.character(Country),
                          "\n(click to open Google search)")
  )

p <- ggplot(data = latest_click, aes(x = Old, y = Young, colour = Continent)) +
  geom_point_interactive(
    aes(tooltip = tooltip_text,
        onclick  = onclick_js),
    size = 3.5, alpha = 0.7
  ) +
  scale_colour_viridis_d(option = "plasma") +
  labs(title = "Click any country point to open a Google search for that country",
       x = "% Aged", y = "% Young") +
  theme_minimal() +
  theme(plot.title = element_text(face = "bold", size = 13))

girafe(
  ggobj      = p,
  width_svg  = 10,
  height_svg = 10 * 0.618,
  options    = list(
    opts_hover(css = "fill:#FF3333; cursor:pointer;")
  )
)
```

Web document link with a data object will be displayed on the web browser upon mouse click.

::: {.callout-warning}
Note that click actions must be a **string column** in the dataset containing valid JavaScript instructions. The `onclick` value is evaluated as raw JavaScript in the browser — errors in the JS string will silently fail.
:::

::: {.callout-note}
## What I Learned
The `onclick` aesthetic turns every data point into a hyperlink — bridging the visualisation to any external information system. `cursor:pointer` in `opts_hover()` signals clickability to users without additional UI elements. Always pre-compute the JS string as a data column (never build it inside `aes()`), and always use `URLencode()` for country names to handle spaces and special characters correctly.
:::

---

### Coordinated Multiple Views with ggiraph {#ggiraph-coordinated}

When a data point of one of the scatter plots is selected, the corresponding data point on the second visualisation will be highlighted too.

In order to build coordinated multiple views, the following programming strategy is used:

1. Appropriate interactive functions of **ggiraph** are used to create the multiple views.
2. `patchwork` function of the **patchwork** package is used *inside* `girafe()` via the `code` argument to create the interactive coordinated multiple views.

```{r}
#| label: ggiraph-coordinated

latest_coord <- pop_data %>%
  filter(Year == max(Year, na.rm = TRUE)) %>%
  mutate(tooltip_text = paste0(
    "<b>", Country, "</b><br>",
    "Continent: ", Continent, "<br>",
    "% Aged: ",   round(Old, 1),   "%<br>",
    "% Young: ",  round(Young, 1), "%"
  ))

p1 <- ggplot(data = latest_coord, aes(x = Old, y = Young, colour = Continent)) +
  geom_point_interactive(
    aes(data_id = Country,
        tooltip = tooltip_text),
    size = 3, alpha = 0.7
  ) +
  scale_colour_viridis_d(option = "plasma") +
  labs(title = "% Aged vs % Young",
       x = "% Aged", y = "% Young") +
  theme_minimal() +
  theme(legend.position = "none",
        plot.title = element_text(face = "bold", size = 11))

p2 <- ggplot(data = latest_coord, aes(x = Old, y = Population / 1e6, colour = Continent)) +
  geom_point_interactive(
    aes(data_id = Country,
        tooltip = tooltip_text),
    size = 3, alpha = 0.7
  ) +
  scale_colour_viridis_d(option = "turbo") +
  labs(title = "% Aged vs Population (Millions)",
       x = "% Aged", y = "Population (M)") +
  theme_minimal() +
  theme(legend.position = "none",
        plot.title = element_text(face = "bold", size = 11))

girafe(
  code       = print(p1 + p2),
  width_svg  = 12,
  height_svg = 4.5,
  options    = list(
    opts_hover(css     = "fill:#202020; stroke:#202020;"),
    opts_hover_inv(css = "opacity:0.2;")
  )
)
```

The `data_id` aesthetic is critical to link observations between plots — here `Country` is the shared key. The tooltip aesthetic is optional but provides useful per-point detail when hovering.

::: {.callout-important}
## What I Learned
Coordinated views require two conditions: (1) identical `data_id` keys across all panels — `Country` here — and (2) the patchwork combination placed *inside* `girafe()` via the `code` argument rather than `ggobj`. When both conditions are met, ggiraph handles all JavaScript linking automatically. This enables a powerful workflow: identify an outlier in the demographic scatter (left panel) and instantly see its population size context (right panel) — something a static chart grid can never achieve.
:::

---

## Interactive Data Visualisation — plotly Methods {#plotly-methods}

::: {.callout-tip}
## Core Concept
**plotly**'s R library creates interactive web graphics via two approaches:

- `plot_ly()` — native plotly syntax using `~` column references
- `ggplotly()` — converts an existing ggplot2 object

Both produce charts with built-in zoom, pan, hover tooltips, and download buttons. plotly renders via JavaScript (`plotly.js`) rather than SVG, making it particularly suited for web dashboards.
:::

### Creating an Interactive Scatter Plot: `plot_ly()` Method {#plotly-basic}

The tabset below shows a basic interactive plot created by using `plot_ly()`:

::: {.panel-tabset}

### The Plot

```{r}
#| code-fold: false
#| echo: false

latest_pl <- pop_data %>% filter(Year == max(Year, na.rm = TRUE))

plot_ly(data = latest_pl,
        x    = ~Old,
        y    = ~Young)
```

### The Code Chunk

```{r}
#| eval: false

latest_pl <- pop_data %>% filter(Year == max(Year, na.rm = TRUE))

plot_ly(data = latest_pl,
        x    = ~Old,
        y    = ~Young)
```

:::

The default output is already interactive: hover to see coordinates, drag to zoom, double-click to reset.

::: {.callout-note}
## What I Learned
`plot_ly()` produces a fully interactive chart with a single function call. The tilde syntax (`~Old`) is distinctive to plotly and signals a column reference, not a literal value. The default toolbar (zoom, pan, box select, lasso, download as PNG) is added automatically and requires no configuration — a stark contrast to the manual `girafe()` setup in ggiraph.
:::

---

### Working with Visual Variables: `plot_ly()` Method {#plotly-color}

The `color` argument maps a qualitative variable to colour. Clicking legend symbols toggles trace visibility:

::: {.panel-tabset}

### The Plot

```{r}
#| code-fold: false
#| echo: false

latest_pl <- pop_data %>% filter(Year == max(Year, na.rm = TRUE))

plot_ly(data   = latest_pl,
        x      = ~Old,
        y      = ~Young,
        color  = ~Continent,
        colors = viridis::viridis(length(unique(latest_pl$Continent))))
```

### The Code Chunk

```{r}
#| eval: false

latest_pl <- pop_data %>% filter(Year == max(Year, na.rm = TRUE))

plot_ly(data   = latest_pl,
        x      = ~Old,
        y      = ~Young,
        color  = ~Continent,
        colors = viridis::viridis(length(unique(latest_pl$Continent))))
```

:::

Click on the colour symbol at the legend to toggle groups on/off. Double-click to isolate one continent.

::: {.callout-note}
## What I Learned
Mapping `color` to a factor automatically creates a legend with toggle interactivity — a feature that would require custom JavaScript in a static chart. For demographic data, clicking continent labels to isolate groups is far more intuitive than filtering data and re-rendering. The `colors` argument accepts a palette function; `viridis::viridis()` generates the correct number of accessible colours automatically without requiring a named vector.
:::

---

### Creating an Interactive Scatter Plot: `ggplotly()` Method {#ggplotly-method}

`ggplotly()` converts any existing ggplot2 object into a plotly chart:

::: {.panel-tabset}

### The Plot

```{r}
#| code-fold: false
#| echo: false

latest_gg <- pop_data %>% filter(Year == max(Year, na.rm = TRUE))

p <- ggplot(data = latest_gg, aes(x = Old, y = Young, colour = Continent)) +
  geom_point(size = 2.5, alpha = 0.7) +
  scale_colour_viridis_d(option = "plasma") +
  labs(title = "% Aged vs % Young",
       x = "% Aged", y = "% Young") +
  theme_minimal()

ggplotly(p)
```

### The Code Chunk

```{r}
#| eval: false

latest_gg <- pop_data %>% filter(Year == max(Year, na.rm = TRUE))

p <- ggplot(data = latest_gg, aes(x = Old, y = Young, colour = Continent)) +
  geom_point(size = 2.5, alpha = 0.7) +
  scale_colour_viridis_d(option = "plasma") +
  labs(title = "% Aged vs % Young",
       x = "% Aged", y = "% Young") +
  theme_minimal()

ggplotly(p)
```

:::

Notice that the only extra line needed is `ggplotly(p)`. The zoom toolbar, hover tooltips, and legend toggles are added automatically.

::: {.callout-note}
## What I Learned
`ggplotly()` is the bridge between the ggplot2 ecosystem and plotly interactivity. When you already have carefully crafted ggplot code with proper themes, colour scales, and labels, `ggplotly()` preserves all of it while adding the full plotly toolbar. The trade-off: some advanced ggplot2 features may not translate perfectly. For standard scatter and bar charts, the conversion is reliable and fast.
:::

---

### Coordinated Multiple Views with plotly {#plotly-coordinated}

The creation of a coordinated linked plot by using plotly involves three steps:

- `highlight_key()` of **plotly** package is used as shared data
- Two scatter plots are created by using ggplot2 functions
- `subplot()` of **plotly** package is used to place them side by side

::: {.panel-tabset}

### The Plot

```{r}
#| code-fold: false
#| echo: false

latest_sub <- pop_data %>% filter(Year == max(Year, na.rm = TRUE))

d <- highlight_key(latest_sub)

p1 <- ggplot(data = d, aes(x = Old, y = Young)) +
  geom_point(size = 2, alpha = 0.7, colour = "#831843") +
  labs(title = "% Aged vs % Young",
       x = "% Aged", y = "% Young") +
  theme_minimal()

p2 <- ggplot(data = d, aes(x = Old, y = Population / 1e6)) +
  geom_point(size = 2, alpha = 0.7, colour = "#1e40af") +
  labs(title = "% Aged vs Population (M)",
       x = "% Aged", y = "Population (Millions)") +
  theme_minimal()

subplot(ggplotly(p1), ggplotly(p2))
```

### The Code Chunk

```{r}
#| eval: false

latest_sub <- pop_data %>% filter(Year == max(Year, na.rm = TRUE))

d <- highlight_key(latest_sub)

p1 <- ggplot(data = d, aes(x = Old, y = Young)) +
  geom_point(size = 2, alpha = 0.7, colour = "#831843") +
  labs(title = "% Aged vs % Young",
       x = "% Aged", y = "% Young") +
  theme_minimal()

p2 <- ggplot(data = d, aes(x = Old, y = Population / 1e6)) +
  geom_point(size = 2, alpha = 0.7, colour = "#1e40af") +
  labs(title = "% Aged vs Population (M)",
       x = "% Aged", y = "Population (Millions)") +
  theme_minimal()

subplot(ggplotly(p1), ggplotly(p2))
```

:::

Click on a data point of one scatter plot and see how the corresponding point on the other is selected.

- `highlight_key()` simply creates an object of class `crosstalk::SharedData`. Visit this [link](https://rstudio.github.io/crosstalk/) to learn more about crosstalk.
- `subplot()` places plotly charts side by side while maintaining the shared data linkage.

::: {.callout-note}
## What I Learned
`highlight_key()` is the glue between panels — it wraps a data frame with a SharedData object so plotly knows which rows correspond across charts. Without it, `subplot()` produces two fully independent charts. The pattern is always: shared data → two ggplots → two `ggplotly()` calls → one `subplot()`. This lets viewers ask multi-variable questions — "Is this highly-aged country also unusually large in population?" — answered with a single click.
:::

---

## Interactive Data Visualisation — crosstalk Methods {#crosstalk-methods}

::: {.callout-tip}
## Core Concept
[crosstalk](https://rstudio.github.io/crosstalk/index.html) is an add-on to the htmlwidgets package. It extends htmlwidgets with a set of classes, functions, and conventions for implementing cross-widget interactions — currently **linked brushing** and **filtering**. It works with plotly, DT, and leaflet.
:::

### Interactive Data Table: DT Package {#dt-table}

The **DT** package is a wrapper of the JavaScript library DataTables. Data objects in R can be rendered as HTML tables using the JavaScript library DataTables (typically via R Markdown or Shiny):

```{r}
#| label: dt-table
DT::datatable(
  pop_data %>% filter(Year == max(Year, na.rm = TRUE)),
  class   = "compact",
  caption = paste("GlobalPopulation data — Year", max(pop_data$Year, na.rm = TRUE)),
  options = list(pageLength = 10, scrollX = TRUE)
)
```

::: {.callout-note}
## What I Learned
`DT::datatable()` is the fastest way to make any R data frame explorable in an HTML report. The search box — filtering hundreds of rows in real time — replaces multiple manual `filter()` calls. Tables and charts are **complementary**: charts show population-level patterns (clusters, trends), tables expose individual country-level values. This complementarity is exactly what linked brushing below exploits.
:::

---

### Linked Brushing: crosstalk Method {#crosstalk-brushing}

::: {.panel-tabset}

### The Plot

```{r}
#| code-fold: false
#| echo: false

latest_brush <- pop_data %>% filter(Year == max(Year, na.rm = TRUE))

d <- highlight_key(latest_brush)

p <- ggplot(d, aes(Old, Young, colour = Continent)) +
  geom_point(size = 2.5, alpha = 0.7) +
  scale_colour_viridis_d(option = "plasma") +
  labs(x = "% Aged", y = "% Young") +
  theme_minimal()

gg <- highlight(ggplotly(p), "plotly_selected")

crosstalk::bscols(
  gg,
  DT::datatable(d, options = list(pageLength = 8, scrollX = TRUE)),
  widths = 5
)
```

### The Code Chunk

```{r}
#| eval: false

latest_brush <- pop_data %>% filter(Year == max(Year, na.rm = TRUE))

d <- highlight_key(latest_brush)

p <- ggplot(d, aes(Old, Young, colour = Continent)) +
  geom_point(size = 2.5, alpha = 0.7) +
  scale_colour_viridis_d(option = "plasma") +
  labs(x = "% Aged", y = "% Young") +
  theme_minimal()

gg <- highlight(ggplotly(p), "plotly_selected")

crosstalk::bscols(
  gg,
  DT::datatable(d, options = list(pageLength = 8, scrollX = TRUE)),
  widths = 5
)
```

:::

Click and drag to select points in the scatter plot — the table on the right filters to show only those rows.

Things to learn from the code:

- `highlight()` is a **plotly** function that sets brushing options. `"plotly_selected"` specifies that selection is triggered by drag-select (lasso or box). These options are designed for linking multiple plotly graphs and may not behave as expected when linking to another htmlwidget via crosstalk.
- `bscols()` is a **crosstalk** helper function that places HTML elements side by side using Bootstrap columns. It is designed for R Markdown / Quarto documents. **Warning:** This will bring in all of Bootstrap!
- `widths = 5` allocates 5/12 Bootstrap columns to each widget (plotly + DT), leaving 2 as margin.

::: {.callout-important}
## What I Learned
Linked brushing with crosstalk combines three layers: plotly handles the *selection event* (`plotly_selected`), `highlight_key()` propagates the selection to all widgets sharing the same SharedData object, and DT reacts by filtering its rows. The result is a powerful *explore-and-drill-down* workflow: spot a demographic cluster in the chart (e.g., a group of highly-aged, low-young countries), drag to select it, and immediately read their exact country names and population figures in the table — all without a single line of JavaScript.

This is the highest-level interactivity achievable in a static Quarto HTML document without Shiny — and for many analytical tasks, it is sufficient.
:::

---

## Comparison: ggiraph vs plotly vs crosstalk {#comparison}

::: {.panel-tabset}

### ggiraph

**Best for:**

- Embedding interactive SVG charts in standalone HTML reports
- Group-level hover highlighting via `data_id`
- Coordinated multiple views within a single `girafe()` call
- Full ggplot2 theme, scale and geom compatibility
- Offline use (no external JS CDN required)

**Key functions:** `geom_*_interactive()`, `girafe()`, `opts_tooltip()`, `opts_hover()`, `opts_hover_inv()`

### plotly

**Best for:**

- Richer built-in toolbar (zoom, pan, box/lasso select, download PNG)
- Interactive legends with click-to-toggle trace visibility
- Coordinated views via `highlight_key()` + `subplot()`
- Animation support (see Exercise 3b)
- Web dashboards and Shiny integration

**Key functions:** `plot_ly()`, `ggplotly()`, `highlight_key()`, `highlight()`, `subplot()`

### crosstalk

**Best for:**

- Linking a chart to a sortable, searchable data table (DT)
- Drill-down exploration: select visual cluster → read exact values
- No Shiny server required — works in any HTML document

**Key functions:** `highlight_key()`, `highlight()`, `bscols()`

### Decision Guide

| Criterion | ggiraph | plotly | crosstalk |
|-----------|---------|--------|-----------|
| Output format | SVG (inline HTML) | JS canvas | HTML layout |
| Tooltip customisation | Full CSS | Layout via `text =` | Via plotly |
| Group hover highlight | ✅ `data_id` | Legend toggle | ✅ via SharedData |
| Linked chart + table | ❌ | ❌ alone | ✅ DT |
| Animation support | ❌ | ✅ `frame =` | ❌ |
| Shiny integration | Partial | Full | Partial |
| Ease for ggplot users | ✅ drop-in | Medium | Medium |

:::

::: {.callout-note}
## Overall What I Learned
The key insight is that interactivity is not decoration — it is a **layer of analysis** that static charts cannot provide.

- **ggiraph** is the most natural upgrade for ggplot2 users: swap `geom_*` → `geom_*_interactive`, add `tooltip`/`data_id`, wrap in `girafe()`. It is ideal when the goal is group-level highlighting across coordinated views within a report.
- **plotly** excels when you need a richer toolbar or plan to extend to animations. `ggplotly()` bridges the ggplot2 workflow with minimal friction.
- **crosstalk** enables the most powerful analytical workflow in a static document: drag-select a visual cluster, instantly read its members in a table — simulating a database query without any server code.

Choosing between them is a question of audience and context: ggiraph for polished reports, plotly for dashboards, crosstalk for exploratory analysis handoffs.
:::

---


# References

## ggiraph

- [ggiraph package documentation](https://davidgohel.github.io/ggiraph/index.html) — online reference guide and articles
- [Customizing girafe objects](https://www.ardata.fr/ggiraph-book/customize.html) — CSS options for tooltip, hover, and selection
- [ggiraph Book — Selections](https://www.ardata.fr/ggiraph-book/selections.html) — data_id, Shiny integration

## plotly for R

- [Getting Started with Plotly in R](https://plotly.com/r/getting-started/)
- Carson Sievert (2020) *Interactive web-based data visualization with R, plotly, and shiny*, Chapman and Hall/CRC — [online version](https://plotly-r.com/)
- [Plotly R Figure Reference](https://plotly.com/r/reference/index/) — comprehensive visual reference
- [Plotly R Library Fundamentals](https://plotly.com/r/plotly-fundamentals/)

## crosstalk & DT

- [crosstalk documentation](https://rstudio.github.io/crosstalk/)
- [DT package guide](https://rstudio.github.io/DT/)

## Source

- [R for Visual Analytics Chapter 3](https://r4va.netlify.app/chap03)