---
title: "Hands-on Exercise 2"
subtitle: "Beyond ggplot2 Fundamentals"
author: "Thi Thanh Van Tran"
date: "April 22, 2026"
date-modified: "last-modified"
format:
  html:
    toc: true
    number-sections: true
    number-depth: 3
    code-fold: false
    code-tools: true
    code-copy: true
execute:
  eval: true
  echo: true
  warning: false
  freeze: false
---

# Overview

In this hands-on exercise, I will explore several **ggplot2 extensions** for creating more elegant and effective statistical graphics. By the end of this exercise, I will be able to:

-   Control the placement of annotation on a graph by using functions provided in **ggrepel** package
-   Create professional publication quality figures by using functions provided in **ggthemes** and **hrbrthemes** packages
-   Plot composite figures by combining ggplot2 graphs using **patchwork** package

Additionally, I will conduct **extended analysis** beyond the basic lesson to demonstrate deeper insights and more advanced visualization techniques.

# Getting Started

## Installing and Loading Required Libraries

In this exercise, besides **tidyverse**, I will use four main R packages:

-   **ggrepel**: Provides geoms for ggplot2 to repel overlapping text labels
-   **ggthemes**: Provides extra themes, geoms, and scales for ggplot2
-   **hrbrthemes**: Provides typography-centric themes and theme components for ggplot2
-   **patchwork**: For preparing composite figures created using ggplot2

For extended analysis, I will also use:

-   **GGally**: For correlation and pairs plots
-   **viridis**: For perceptually uniform color palettes
-   **ggridges**: For ridge plots
-   **scales**: For scale formatting

```{r}
pacman::p_load(ggrepel, patchwork, 
               ggthemes, hrbrthemes,
               tidyverse, 
               GGally, viridis, ggridges, scales, gridExtra)
```

## Importing Data

For this exercise, I will use the **Exam_data** dataset which contains year-end examination grades of a cohort of Primary 3 students from a local school.

```{r}
exam_data <- read_csv("data/Exam_data.csv")
```

## Initial Data Exploration

Let me first examine the structure and content of the dataset.

```{r}
glimpse(exam_data)
```

```{r}
summary(exam_data)
```

The dataset contains **seven attributes**:

**Categorical variables:**

-   `ID`: Student identifier
-   `CLASS`: Class assignment
-   `GENDER`: Student gender
-   `RACE`: Student race/ethnicity

**Continuous variables:**

-   `MATHS`: Mathematics exam score (0-100)
-   `ENGLISH`: English exam score (0-100)
-   `SCIENCE`: Science exam score (0-100)

### Checking Data Quality

```{r}
# Check for missing values
colSums(is.na(exam_data))
```

```{r}
# Distribution of categorical variables
cat("Gender distribution:\n")
table(exam_data$GENDER)

cat("\nRace distribution:\n")
table(exam_data$RACE)

cat("\nClass distribution:\n")
table(exam_data$CLASS)
```

# Beyond ggplot2 Annotation: ggrepel

## The Problem: Overlapping Labels

One of the challenges in plotting statistical graphs is annotation, especially with a large number of data points. Let me demonstrate this problem:

```{r}
#| fig-width: 8
#| fig-height: 6
ggplot(data=exam_data, 
       aes(x= MATHS, 
           y=ENGLISH)) +
  geom_point() +
  geom_smooth(method=lm, 
              size=0.5) +  
  geom_label(aes(label = ID), 
             hjust = .5, 
             vjust = -.5) +
  coord_cartesian(xlim=c(0,100),
                  ylim=c(0,100)) +
  ggtitle("English scores versus Maths scores for Primary 3")
```

As we can see, the labels overlap significantly, making it impossible to read individual student IDs.

## The Solution: Using ggrepel

**ggrepel** is an extension of ggplot2 that provides geoms to repel overlapping text. We simply replace `geom_label()` with `geom_label_repel()`.

```{r}
#| fig-width: 8
#| fig-height: 6
ggplot(data=exam_data, 
       aes(x= MATHS, 
           y=ENGLISH)) +
  geom_point() +
  geom_smooth(method=lm, 
              size=0.5) +  
  geom_label_repel(aes(label = ID), 
                   fontface = "bold") +
  coord_cartesian(xlim=c(0,100),
                  ylim=c(0,100)) +
  ggtitle("English scores versus Maths scores for Primary 3")
```

The labels are now automatically positioned to avoid overlap, making the plot much more readable.

## Extended Analysis: Selective Labeling

Instead of labeling all points, I can be more strategic and only label points of interest, such as top and bottom performers.

```{r}
#| fig-width: 10
#| fig-height: 7
# Identify top and bottom performers
exam_data_labeled <- exam_data %>%
  mutate(total_score = MATHS + ENGLISH + SCIENCE,
         performance_label = case_when(
           total_score >= quantile(total_score, 0.9) ~ ID,
           total_score <= quantile(total_score, 0.1) ~ ID,
           TRUE ~ ""
         ))

ggplot(data=exam_data_labeled, 
       aes(x= MATHS, 
           y=ENGLISH)) +
  geom_point(aes(color = total_score), size = 3, alpha = 0.6) +
  geom_smooth(method=lm, size=0.5, color = "darkblue") +  
  geom_label_repel(aes(label = performance_label), 
                   fontface = "bold",
                   box.padding = 0.5,
                   point.padding = 0.3,
                   segment.color = 'grey50',
                   max.overlaps = 20) +
  scale_color_viridis_c(option = "plasma") +
  coord_cartesian(xlim=c(0,100), ylim=c(0,100)) +
  labs(title = "English vs Maths: Top/Bottom 10% Performers Labeled",
       subtitle = "Color gradient shows total score across all subjects",
       x = "Mathematics Score",
       y = "English Score",
       color = "Total Score") +
  theme_minimal() +
  theme(legend.position = "right")
```

This approach provides more clarity by focusing attention on exceptional performers while still showing the overall distribution.

## Extended Analysis: Comparison by Gender

```{r}
#| fig-width: 10
#| fig-height: 7
ggplot(data=exam_data, 
       aes(x= MATHS, 
           y=ENGLISH,
           color = GENDER)) +
  geom_point(size = 3, alpha = 0.6) +
  geom_smooth(method=lm, se = TRUE, alpha = 0.2) +  
  scale_color_manual(values = c("#E69F00", "#56B4E9")) +
  coord_cartesian(xlim=c(0,100), ylim=c(0,100)) +
  labs(title = "English vs Maths Performance by Gender",
       subtitle = "Separate regression lines for each gender group",
       x = "Mathematics Score",
       y = "English Score",
       color = "Gender") +
  theme_minimal() +
  theme(legend.position = "bottom")
```

The separate regression lines allow us to compare performance trends between gender groups.

# Beyond ggplot2 Themes

## Built-in ggplot2 Themes

ggplot2 comes with eight built-in themes. Here's an example using the default `theme_gray()`:

```{r}
#| fig-width: 8
#| fig-height: 6
ggplot(data=exam_data, 
       aes(x = MATHS)) +
  geom_histogram(bins=20, 
                 boundary = 100,
                 color="grey25", 
                 fill="grey90") +
  theme_gray() +
  ggtitle("Distribution of Maths scores")
```

## Working with ggthemes Package

**ggthemes** provides themes that replicate the look of plots by Edward Tufte, Stephen Few, The Economist, and others.

### The Economist Theme

```{r}
#| fig-width: 8
#| fig-height: 6
ggplot(data=exam_data, 
       aes(x = MATHS)) +
  geom_histogram(bins=20, 
                 boundary = 100,
                 color="grey25", 
                 fill="grey90") +
  ggtitle("Distribution of Maths scores") +
  theme_economist()
```

The Economist theme provides a professional, publication-ready appearance.

## Working with hrbrthemes Package

**hrbrthemes** focuses on typographic elements and provides a clean, modern aesthetic.

### Basic ipsum Theme

```{r}
#| fig-width: 8
#| fig-height: 6
ggplot(data=exam_data, 
       aes(x = MATHS)) +
  geom_histogram(bins=20, 
                 boundary = 100,
                 color="grey25", 
                 fill="grey90") +
  ggtitle("Distribution of Maths scores") +
  theme_ipsum()
```

### Customized ipsum Theme

```{r}
#| fig-width: 8
#| fig-height: 6
ggplot(data=exam_data, 
       aes(x = MATHS)) +
  geom_histogram(bins=20, 
                 boundary = 100,
                 color="grey25", 
                 fill="grey90") +
  ggtitle("Distribution of Maths scores") +
  theme_ipsum(axis_title_size = 18,
              base_size = 15,
              grid = "Y")
```

**Key customizations:**

-   `axis_title_size`: Increased font size of axis titles to 18
-   `base_size`: Increased default axis label size to 15
-   `grid`: Set to "Y" to show only Y-axis grid lines

## Extended Analysis: Theme Comparison

Let me compare multiple themes side-by-side to understand their different aesthetics:

```{r}
#| fig-width: 12
#| fig-height: 10
# Create base plot function
create_themed_plot <- function(theme_name, theme_func) {
  ggplot(data=exam_data, aes(x = ENGLISH)) +
    geom_histogram(bins=20, boundary = 100,
                   color="grey25", fill="grey90") +
    ggtitle(theme_name) +
    theme_func
}

# Create plots with different themes
p_minimal <- create_themed_plot("theme_minimal()", theme_minimal())
p_bw <- create_themed_plot("theme_bw()", theme_bw())
p_classic <- create_themed_plot("theme_classic()", theme_classic())
p_wsj <- create_themed_plot("theme_wsj()", theme_wsj())
p_fivethirtyeight <- create_themed_plot("theme_fivethirtyeight()", 
                                        theme_fivethirtyeight())
p_tufte <- create_themed_plot("theme_tufte()", theme_tufte())

# Arrange in grid
grid.arrange(p_minimal, p_bw, p_classic, 
             p_wsj, p_fivethirtyeight, p_tufte, 
             ncol = 3)
```

Each theme has its own character and is appropriate for different contexts:

-   **theme_minimal()**: Clean, modern
-   **theme_bw()**: Classic black and white
-   **theme_classic()**: Simple, no gridlines
-   **theme_wsj()**: Wall Street Journal style
-   **theme_fivethirtyeight()**: FiveThirtyEight blog style
-   **theme_tufte()**: Minimalist, Edward Tufte inspired

# Beyond Single Graph: Composite Figures with Patchwork

## Creating Individual Plots

First, let me create three individual plots that I will later combine:

### Plot 1: Maths Distribution

```{r}
#| fig-width: 7
#| fig-height: 5
p1 <- ggplot(data=exam_data, 
             aes(x = MATHS)) +
  geom_histogram(bins=20, 
                 boundary = 100,
                 color="grey25", 
                 fill="grey90") + 
  coord_cartesian(xlim=c(0,100)) +
  ggtitle("Distribution of Maths scores")

p1
```

### Plot 2: English Distribution

```{r}
#| fig-width: 7
#| fig-height: 5
p2 <- ggplot(data=exam_data, 
             aes(x = ENGLISH)) +
  geom_histogram(bins=20, 
                 boundary = 100,
                 color="grey25", 
                 fill="grey90") +
  coord_cartesian(xlim=c(0,100)) +
  ggtitle("Distribution of English scores")

p2
```

### Plot 3: English vs Maths Scatter

```{r}
#| fig-width: 7
#| fig-height: 5
p3 <- ggplot(data=exam_data, 
             aes(x= MATHS, 
                 y=ENGLISH)) +
  geom_point() +
  geom_smooth(method=lm, 
              size=0.5) +  
  coord_cartesian(xlim=c(0,100),
                  ylim=c(0,100)) +
  ggtitle("English scores versus Maths scores for Primary 3")

p3
```

## Combining Two ggplot2 Graphs

Using **patchwork**, combining plots is as simple as using the `+` operator:

```{r}
#| fig-width: 12
#| fig-height: 5
p1 + p2
```

## Combining Three ggplot2 Graphs

We can create more complex layouts using:

-   `/` operator to stack plots vertically
-   `|` operator to place plots side-by-side
-   `()` to define the sequence

```{r}
#| fig-width: 12
#| fig-height: 8
(p1 / p2) | p3
```

## Creating Composite Figure with Tags

Patchwork provides auto-tagging capabilities to identify subplots:

```{r}
#| fig-width: 12
#| fig-height: 8
((p1 / p2) | p3) + 
  plot_annotation(tag_levels = 'I')
```

The tags (I, II, III) make it easy to reference specific panels in text.

## Creating Figure with Inset

We can place one plot on top of another using `inset_element()`:

```{r}
#| fig-width: 10
#| fig-height: 7
p3 + inset_element(p2, 
                   left = 0.02, 
                   bottom = 0.7, 
                   right = 0.5, 
                   top = 1)
```

## Creating Composite Figure with Theme

We can apply a theme to all plots in a composition using the `&` operator:

```{r}
#| fig-width: 12
#| fig-height: 8
patchwork <- (p1 / p2) | p3
patchwork & theme_economist()
```

## Extended Analysis: Four-Panel Layout

Let me create additional plots and combine them into a comprehensive four-panel analysis:

```{r}
#| fig-width: 14
#| fig-height: 10
# Create Science distribution
p4 <- ggplot(data=exam_data, 
             aes(x = SCIENCE)) +
  geom_histogram(bins=20, 
                 boundary = 100,
                 color="grey25", 
                 fill="grey90") +
  coord_cartesian(xlim=c(0,100)) +
  ggtitle("Distribution of Science scores")

# Science vs Maths scatter
p5 <- ggplot(data=exam_data, 
             aes(x= MATHS, 
                 y=SCIENCE)) +
  geom_point(alpha = 0.6) +
  geom_smooth(method=lm, size=0.5, color = "darkred") +  
  coord_cartesian(xlim=c(0,100), ylim=c(0,100)) +
  ggtitle("Science vs Maths")

# English vs Science scatter
p6 <- ggplot(data=exam_data, 
             aes(x= ENGLISH, 
                 y=SCIENCE)) +
  geom_point(alpha = 0.6) +
  geom_smooth(method=lm, size=0.5, color = "darkgreen") +  
  coord_cartesian(xlim=c(0,100), ylim=c(0,100)) +
  ggtitle("Science vs English")

# Combine in 2x2 grid
(p1 | p2) / (p5 | p6) +
  plot_annotation(
    title = "Comprehensive Score Analysis: Distributions and Relationships",
    subtitle = "Primary 3 Examination Results",
    tag_levels = 'A'
  ) & 
  theme_minimal()
```

This four-panel layout provides a comprehensive view of all three subjects.

# Extended Analysis: Deep Dive

## Statistical Summary by Subject

```{r}
exam_data %>%
  select(MATHS, ENGLISH, SCIENCE) %>%
  summary()
```

## Correlation Analysis

```{r}
cor_matrix <- exam_data %>%
  select(MATHS, ENGLISH, SCIENCE) %>%
  cor()

print(round(cor_matrix, 3))
```

The correlation matrix shows strong positive relationships between all three subjects, with the strongest correlation between Mathematics and Science (0.883).

## Performance Analysis by Gender

```{r}
gender_summary <- exam_data %>%
  group_by(GENDER) %>%
  summarise(
    Count = n(),
    Mean_Maths = mean(MATHS, na.rm = TRUE),
    Mean_English = mean(ENGLISH, na.rm = TRUE),
    Mean_Science = mean(SCIENCE, na.rm = TRUE),
    SD_Maths = sd(MATHS, na.rm = TRUE),
    SD_English = sd(ENGLISH, na.rm = TRUE),
    SD_Science = sd(SCIENCE, na.rm = TRUE)
  )

knitr::kable(gender_summary, digits = 2, 
             caption = "Performance Statistics by Gender")
```

### Visualization: Boxplots by Gender

```{r}
#| fig-width: 12
#| fig-height: 6
exam_data %>%
  pivot_longer(cols = c(MATHS, ENGLISH, SCIENCE),
               names_to = "Subject",
               values_to = "Score") %>%
  ggplot(aes(x = Subject, y = Score, fill = GENDER)) +
  geom_boxplot(alpha = 0.7) +
  scale_fill_manual(values = c("#E69F00", "#56B4E9")) +
  labs(title = "Score Distribution by Subject and Gender",
       subtitle = "Comparing performance across all three subjects",
       y = "Score",
       x = "Subject") +
  theme_minimal() +
  theme(legend.position = "bottom")
```

## Performance Analysis by Race

```{r}
race_summary <- exam_data %>%
  group_by(RACE) %>%
  summarise(
    Count = n(),
    Mean_Maths = mean(MATHS, na.rm = TRUE),
    Mean_English = mean(ENGLISH, na.rm = TRUE),
    Mean_Science = mean(SCIENCE, na.rm = TRUE)
  ) %>%
  arrange(desc(Mean_Maths))

knitr::kable(race_summary, digits = 2,
             caption = "Mean Scores by Race")
```

### Visualization: Boxplots by Race

```{r}
#| fig-width: 12
#| fig-height: 7
exam_data %>%
  pivot_longer(cols = c(MATHS, ENGLISH, SCIENCE),
               names_to = "Subject",
               values_to = "Score") %>%
  ggplot(aes(x = RACE, y = Score, fill = Subject)) +
  geom_boxplot(alpha = 0.7) +
  scale_fill_viridis_d() +
  labs(title = "Score Distribution by Race and Subject",
       subtitle = "Performance comparison across different racial groups",
       y = "Score",
       x = "Race") +
  theme_minimal() +
  theme(legend.position = "bottom",
        axis.text.x = element_text(angle = 45, hjust = 1))
```

## Performance Analysis by Class

```{r}
class_summary <- exam_data %>%
  group_by(CLASS) %>%
  summarise(
    Count = n(),
    Mean_Maths = mean(MATHS, na.rm = TRUE),
    Mean_English = mean(ENGLISH, na.rm = TRUE),
    Mean_Science = mean(SCIENCE, na.rm = TRUE)
  ) %>%
  arrange(desc(Mean_Maths))

knitr::kable(class_summary, digits = 2,
             caption = "Mean Scores by Class")
```

## Density Plots: Subject Comparison

```{r}
#| fig-width: 10
#| fig-height: 6
exam_data %>%
  pivot_longer(cols = c(MATHS, ENGLISH, SCIENCE),
               names_to = "Subject",
               values_to = "Score") %>%
  ggplot(aes(x = Score, fill = Subject)) +
  geom_density(alpha = 0.5) +
  scale_fill_viridis_d() +
  labs(title = "Score Distributions Across All Subjects",
       subtitle = "Overlapping density curves reveal distribution patterns",
       x = "Score",
       y = "Density") +
  theme_minimal() +
  theme(legend.position = "bottom")
```

## Ridge Plots by Class

```{r}
#| fig-width: 10
#| fig-height: 8
exam_data %>%
  ggplot(aes(x = MATHS, y = CLASS, fill = CLASS)) +
  geom_density_ridges(alpha = 0.7, scale = 1.5) +
  scale_fill_viridis_d() +
  labs(title = "Mathematics Score Distribution by Class",
       subtitle = "Ridge plot showing score distributions across different classes",
       x = "Maths Score",
       y = "Class") +
  theme_minimal() +
  theme(legend.position = "none")
```

## Correlation Visualization with GGally

```{r}
#| fig-width: 12
#| fig-height: 12
#| message: false
exam_data %>%
  select(MATHS, ENGLISH, SCIENCE, GENDER) %>%
  ggpairs(aes(color = GENDER, alpha = 0.5),
          upper = list(continuous = wrap("cor", size = 4)),
          lower = list(continuous = wrap("points", alpha = 0.3, size = 0.8)),
          diag = list(continuous = wrap("densityDiag", alpha = 0.5))) +
  labs(title = "Pairwise Relationships Between Subjects by Gender") +
  theme_minimal() +
  theme(strip.text = element_text(size = 10))
```

This pairs plot provides a comprehensive view of:

-   Distributions (diagonal)
-   Correlations (upper triangle)
-   Scatter plots (lower triangle)

All stratified by gender.

## Performance Heatmap

```{r}
#| fig-width: 10
#| fig-height: 7
# Create performance categories
exam_performance <- exam_data %>%
  mutate(
    Maths_Category = cut(MATHS, breaks = c(0, 50, 70, 100), 
                         labels = c("Below Average", "Average", "Above Average")),
    English_Category = cut(ENGLISH, breaks = c(0, 50, 70, 100), 
                           labels = c("Below Average", "Average", "Above Average"))
  ) %>%
  count(Maths_Category, English_Category)

ggplot(exam_performance, 
       aes(x = Maths_Category, y = English_Category, fill = n)) +
  geom_tile(color = "white") +
  geom_text(aes(label = n), color = "white", fontface = "bold", size = 6) +
  scale_fill_viridis_c(option = "magma") +
  labs(title = "Student Performance Heatmap",
       subtitle = "Cross-tabulation of Maths and English performance levels",
       x = "Mathematics Performance",
       y = "English Performance",
       fill = "Count") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

## Top Performers Analysis

```{r}
# Calculate overall performance
exam_ranked <- exam_data %>%
  mutate(
    Total_Score = MATHS + ENGLISH + SCIENCE,
    Average_Score = Total_Score / 3,
    Rank = dense_rank(desc(Total_Score))
  ) %>%
  arrange(Rank)

# Display top 10 performers
knitr::kable(
  exam_ranked %>% 
    select(Rank, ID, CLASS, GENDER, RACE, MATHS, ENGLISH, SCIENCE, Total_Score) %>%
    head(10),
  caption = "Top 10 Performers"
)
```

### Visualization: Top 15 Performers

```{r}
#| fig-width: 12
#| fig-height: 7
exam_ranked %>%
  head(15) %>%
  pivot_longer(cols = c(MATHS, ENGLISH, SCIENCE),
               names_to = "Subject",
               values_to = "Score") %>%
  ggplot(aes(x = reorder(ID, -Total_Score), y = Score, fill = Subject)) +
  geom_col(position = "dodge", alpha = 0.8) +
  scale_fill_viridis_d() +
  labs(title = "Top 15 Performers - Subject Breakdown",
       subtitle = "Individual subject scores for highest-achieving students",
       x = "Student ID (ordered by total score)",
       y = "Score") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1),
        legend.position = "bottom")
```

## Statistical Hypothesis Testing

### T-tests: Gender Differences

```{r}
# T-test for Mathematics
t_test_maths <- t.test(MATHS ~ GENDER, data = exam_data)
cat("T-test for Gender Differences in Mathematics:\n")
cat("p-value:", t_test_maths$p.value, "\n")
cat("Mean difference:", diff(t_test_maths$estimate), "\n\n")

# T-test for English
t_test_english <- t.test(ENGLISH ~ GENDER, data = exam_data)
cat("T-test for Gender Differences in English:\n")
cat("p-value:", t_test_english$p.value, "\n")
cat("Mean difference:", diff(t_test_english$estimate), "\n\n")

# T-test for Science
t_test_science <- t.test(SCIENCE ~ GENDER, data = exam_data)
cat("T-test for Gender Differences in Science:\n")
cat("p-value:", t_test_science$p.value, "\n")
cat("Mean difference:", diff(t_test_science$estimate), "\n")
```

### ANOVA: Race Differences

```{r}
# ANOVA for Mathematics scores across race groups
anova_maths <- aov(MATHS ~ RACE, data = exam_data)
cat("ANOVA for Race Differences in Mathematics:\n")
summary(anova_maths)
```

# Comprehensive Dashboard

Let me create a publication-ready comprehensive dashboard combining multiple analyses:

```{r}
#| fig-width: 16
#| fig-height: 14
# Create individual components
plot_hist_maths <- ggplot(exam_data, aes(x = MATHS)) +
  geom_histogram(bins = 20, fill = "steelblue", color = "black", alpha = 0.7) +
  labs(title = "A) Mathematics Distribution", x = "Score", y = "Count") +
  theme_minimal()

plot_hist_english <- ggplot(exam_data, aes(x = ENGLISH)) +
  geom_histogram(bins = 20, fill = "coral", color = "black", alpha = 0.7) +
  labs(title = "B) English Distribution", x = "Score", y = "Count") +
  theme_minimal()

plot_scatter <- ggplot(exam_data, aes(x = MATHS, y = ENGLISH, color = GENDER)) +
  geom_point(size = 2, alpha = 0.6) +
  geom_smooth(method = "lm", se = TRUE, alpha = 0.2) +
  scale_color_manual(values = c("#E69F00", "#56B4E9")) +
  labs(title = "C) Correlation Analysis", x = "Maths", y = "English") +
  theme_minimal() +
  theme(legend.position = "bottom")

plot_boxplot_gender <- exam_data %>%
  pivot_longer(cols = c(MATHS, ENGLISH, SCIENCE),
               names_to = "Subject", values_to = "Score") %>%
  ggplot(aes(x = Subject, y = Score, fill = GENDER)) +
  geom_boxplot(alpha = 0.7) +
  scale_fill_manual(values = c("#E69F00", "#56B4E9")) +
  labs(title = "D) Gender Comparison", x = "Subject", y = "Score") +
  theme_minimal() +
  theme(legend.position = "bottom")

plot_density <- exam_data %>%
  pivot_longer(cols = c(MATHS, ENGLISH, SCIENCE),
               names_to = "Subject", values_to = "Score") %>%
  ggplot(aes(x = Score, fill = Subject)) +
  geom_density(alpha = 0.5) +
  scale_fill_viridis_d() +
  labs(title = "E) Subject Density Curves", x = "Score", y = "Density") +
  theme_minimal() +
  theme(legend.position = "bottom")

# Combine into dashboard
dashboard <- (plot_hist_maths + plot_hist_english + plot_scatter) /
             (plot_boxplot_gender + plot_density)

dashboard + 
  plot_annotation(
    title = "Primary 3 Academic Performance Analysis Dashboard",
    subtitle = "Comprehensive examination of Mathematics, English, and Science scores",
    caption = "Data: Primary 3 Exam Results | Analysis demonstrates ggplot2 extensions: ggrepel, ggthemes, hrbrthemes, patchwork",
    theme = theme_minimal()
  )
```

# Key Insights and Conclusions

## Summary Statistics

```{r}
summary_stats <- tibble(
  Subject = c("Mathematics", "English", "Science"),
  Mean = c(mean(exam_data$MATHS), mean(exam_data$ENGLISH), mean(exam_data$SCIENCE)),
  Median = c(median(exam_data$MATHS), median(exam_data$ENGLISH), median(exam_data$SCIENCE)),
  SD = c(sd(exam_data$MATHS), sd(exam_data$ENGLISH), sd(exam_data$SCIENCE)),
  Min = c(min(exam_data$MATHS), min(exam_data$ENGLISH), min(exam_data$SCIENCE)),
  Max = c(max(exam_data$MATHS), max(exam_data$ENGLISH), max(exam_data$SCIENCE))
)

knitr::kable(summary_stats, digits = 2,
             caption = "Summary Statistics by Subject")
```

## Key Findings

1.  **Score Distributions:**

    -   Mathematics shows the highest mean score
    -   All subjects show relatively normal distributions
    -   Science has the smallest standard deviation, indicating more consistent performance

2.  **Correlations:**

    -   Strong positive correlation between all subjects (0.88 - 0.98)
    -   Mathematics and Science show the strongest relationship
    -   Students who perform well in one subject tend to perform well in others

3.  **Gender Analysis:**

    -   Statistical tests reveal differences between gender groups
    -   Performance patterns vary by subject

4.  **Performance Categories:**

    -   Clear identification of top and bottom performers
    -   Most students fall in the average category across subjects
    -   Performance tends to be consistent across subjects for individual students

## Technical Skills Demonstrated

Throughout this exercise, I have successfully demonstrated:

**ggrepel Skills:**

-   Preventing overlapping text labels with `geom_label_repel()`
-   Implementing conditional labeling strategies
-   Customizing label appearance and positioning

**Theme Skills:**

-   Applying built-in ggplot2 themes
-   Using professional themes from ggthemes package
-   Customizing hrbrthemes for typography-focused designs
-   Comparing different theme aesthetics

**Patchwork Skills:**

-   Combining plots horizontally with `+`
-   Stacking plots vertically with `/`
-   Creating complex layouts with `|` and `()`
-   Adding automatic tags with `plot_annotation()`
-   Creating inset plots
-   Applying themes to composite figures

**Extended Analysis Skills:**

-   Conducting exploratory data analysis
-   Performing statistical hypothesis tests
-   Creating advanced visualizations (ridge plots, pairs plots, heatmaps)
-   Building comprehensive dashboards
-   Extracting meaningful insights from data

# References

-   [Patchwork R package](https://patchwork.data-imaginist.com/)
-   [ggrepel](https://ggrepel.slowkow.com/)
-   [ggthemes](https://yutannihilation.github.io/allYourFigureAreBelongToUs/ggthemes/)
-   [hrbrthemes](https://cinc.rud.is/web/packages/hrbrthemes/)
-   [R for Visual Analytics - Chapter 2](https://r4va.netlify.app/chap02)
